---
title: "c9w1 quiz"
author: "Robert Handsfield"
date: "05/10/2015"
output: html_document
---


# Question 1
Consider the following code for the cars data set

```{r}
library(manipulate)
myPlot <- function(s) {
    plot(cars$dist - mean(cars$dist), cars$speed - mean(cars$speed))
    abline(0, s)
}
```

This function plots distance versus speed, each de-meaned and an associated line of slope s. Which of the following code will make a manipulate plot that creates a slider for the slope?

a) __manipulate(myPlot(s), s = slider(0, 2, step = 0.1))__
b) manipulate(myPlot(s), x.s = slider(0, 2, step = 0.1))
c) manipulate(myPlot, s = slider(0, 2, step = 0.1))
d) manipulate(myPlot(s), slider = x(0, 2, step = 0.1))

## Solution

__Answer:__

# Question 2
Which of the following code uses the rCharts package to create a sortable and searchable data table for the airquality data set? Assume the rCharts package and the airquality data set have already been loaded into R.

a) __dTable(airquality, sPaginationType = "full_numbers")__
b) d <- data.frame(airquality, stringsAsFactors = FALSE) print(d)
c) airquality
d) head(airquality)

## Solution

__Answer:__

# Question 3
A basic shiny data product requires:

a) A shiny.R file.
b) __A ui.R and server.R file or a A server.R file and a directory called www containing the relevant html files__
c) A server.R file only.
d) A ui.R file only.


## Solution

__Answer:__

# Question 4
What is incorrect about the followig syntax in ui.R?

```{r}
library(shiny)
shinyUI(pageWithSidebar(
  headerPanel("Data science FTW!"),
  sidebarPanel(
    h2('Big text')
    h3('Sidebar')
  ),
  mainPanel(
      h3('Main Panel text')
  )
))
```

a) The "Sidebar" should say "Sidebar text".
b) The h3 command should be an h2 command.
c) The h2 command has the wrong arguments.
d) __Missing a comma in the sidebar panel__

## Solution

__Answer:__

# Question 5
Consider the following code in ui.R

```{r}
shinyUI(pageWithSidebar(
  headerPanel("Example plot"),
  sidebarPanel(
    sliderInput('mu', 'Guess at the mu',value = 70, min = 60, max = 80, step = 0.05,)
  ),
  mainPanel(
    plotOutput('newHist')
  )
))
```

And the following server.R code.

```{r}
library(UsingR)
data(galton)

shinyServer(
  function(input, output) {
    output$myHist <- renderPlot({
      hist(galton$child, xlab='child height', col='lightblue',main='Histogram')
      mu <- input$mu
      lines(c(mu, mu), c(0, 200),col="red",lwd=5)
      mse <- mean((galton$child - mu)^2)
      text(63, 150, paste("mu = ", mu))
      text(63, 140, paste("MSE = ", round(mse, 2)))
      })
    
  }
)
```



Why isn't it doing what we want?

a) The phrase "Guess at the mu value" should say "mean" instead of "mu"
b) __The server.R output name isn't the same as the plotOutput command used in ui.R.__
c) The limits of the slider are set incorrectly and giving an error.
d) It should be mu <- input$mean in server.R


