---
title: "c9w2 Quiz 2"
author: "Robert Handsfield"
date: "11/8/2015"
output: html_document
---


# Question 1

In the slidify YAML text. Changing the framework from `io2012` to `shower` does what?

a) __It changes the html5 framework thus changing the style of the slides.__  
b) It changes nothing.  
c) It changes whether the document is self contained or requires being connected to the internet.  
d) It changes the ability to display mathjax javascript rendering.  

## Solution:


## Answer:
__Answer:__ 

---

# Question 2

You wrote R code in a slidify document as follows

```{r eval=FALSE}
fit <- lm(y ~ x1 + x2 + x3)
summary(fit)
```

If you want to hide the results of the summary statement (yet still have it run) what should you do?

a) Add a `echo = FALSE` option in the `{r}` call of the code chunk  
b) Comment out the command # summary(fit)  
c) __Add a `results = 'hide'` option in the `{r}` call of the code chunk__  
d) Comment the command as below but also use a bang symbol after the comment, as in `#! summary(fit)`  

## Solution:


## Answer:
__Answer:__ 

---

# Question 3

You wrote R code in a slidify document as follows

```{r eval=FALSE}
fit <- lm(y ~ x1 + x2 + x3)
summary(fit)
```

If you want to display the results, but not the actual code, what should you do?

a) __Add a `echo = FALSE`option in the `{r}` call of the code chunk__  
b) Add a `echo = TRUE` option in the `{r}` call of the code chunk  
c) Comment out the command `# summary(fit)`  
d) Comment the command, but use a bang symbol after the comment, as in `#! summary(fit)`  

## Solution:


## Answer:
__Answer:__ 

---

# Question 4

R studio presentation tool does what?

a) Creates a power point presentation from a generalized markdown format having an extension Rpres.  
b) Creates presentable R code from within presentations. However, it does not actually run the code.  
c) Creates a presentation that can only be run from within Rstudio.  
d) __Creates HTML5 slides using a generalized markdown format having an extention Rpres and creates reproducible presentations by embedding and running the R code from within the presentation document.__  

## Solution:


## Answer:
__Answer:__ 

---

# Question 5

In Rstudio presenter, if you do not want the code to be evaluated, what option do you need to add to the `{r}` options?

a) `run = FALSE`  
b) `echo = FALSE`  
c) __`eval = FALSE`__  
d) `results = 'hide'`  

## Solution:


## Answer:
__Answer:__ 

---

# Question 6

When presenting data analysis to a broad audience, which of the following should be done?

a) Present results in the chronological order in which it was performed.  
b) Show every analysis and method that was done.  
c) Do not include figure captions.  
d) __Summarize the problem first.__  

## Solution:


## Answer:
__Answer:__ 

---
