---
title: "c9w4 Quiz 3"
author: "Robert Handsfield"
date: "11/8/2015"
output: html_document
---


# Question 1

Which of the following items is required for an R package to pass R CMD check without any warnings or errors?

a) vignette
b) a demo directory
c) __An explicit software license__
d) unit tests
e) __DESCRIPTION file__


## Solution


## Answer
__Answer: DESCRIPTION file including a software license__


---

# Question 2

Which of the following is a generic function in a fresh installation of R, with only the default packages loaded?

a) lm  
b) dgamma  
c) colSums
d) __predict__
e) __mean__
f) __show__

## Solution
Investigate each function with `?`.  `?mean` and `?predict` include the phrase
"_generic function_" in their respective Descriptions.

```{r}
?mean
```


## Answer
__Answer: mean__

---

# Question 3

What function is used to obtain the function body for an S4 method function?

a) getS3method()
b) getClass()
c) __getMethod()__
d) showMethods()


## Solution
`methods::getMethod()` displays source code for S4 methods.  
`utils::getS3method()` displays source code for S3 methods.  

## Answer
__Answer: getMethod()__

---

# Question 4

Which one of the following functions must be defined in order to deploy an R function on yhat?

a) model.load
b) model.transform
c) model.predict
d) model.require


## Solution


## Answer


---

# Question 5

Please download the R package DDPQuiz3 from the course web site. Examine the createmean function implemented in the R/ sub-directory. What is the appropriate text to place above the createmean function for Roxygen2 to create a complete help file?

a)
```{r}
#' This function calculates the mean  
#'   
#' @param x is a numeric vector  
#' @return the mean of x  
#' @export  
#' @examples   
#' x <- 1:10  
#' createmean(x)  

```


b)  
```{r}
#' This function calculates the mean  
#'   
#' @param x is a numeric vector  
#' @export  
#' @examples   
#' x <- 1:10  
#' createmean(x)
```  
  
c)  
```{r}
#' This function calculates the mean  
#'   
#' @param x is a numeric vector  
#' @return the mean of x  
#' @export  
#' @examples   
#' x <- 1:10  
#' createmean(y)  
```


d)  
```{r}
#' This function calculates the mean  
#'  
#' @param x is a numeric vector  
#' @return the mean of x  
#' @export  
#' @examples   
#' x <- 1:10  
#' createmean(x)
```


## Solution
For `roxygen2` to create a help file, the following directives are required:

1. `@param` for every input parameter
2. `@return` for the return object/ value
3. `@export` if the function will be exposed to the public API (~public method)
	* this is usually the case, so include the export directive

Optional, but nice

4. `@examples` with __correct syntax__

In the given answer choices, B and D are identical, except that in examples, B stores a sequence in variable _x_, but then calls the function on input _y_.

## Answer
__Answer: D__

