---
title: "R Markdown Demonstration"
author: "Robert Handsfield"
date: "01/18/2015"
output:
  html_document:
    theme: cerulean

---

My First R Markdown File
=========================

This is my first R markdown file.

Load some data here.
```{r}
library(datasets);
data(airquality);
summary(airquality);
```

Make a pairs plot of the data.
```{r}
pairs(airquality)
```
Here's a regression model of ozone on wind, solar radiation, and temp.
```{r}
library(stats)
fit <- lm(Ozone ~ Wind + Solar.R + Temp, data=airquality);
summary(fit);
```

Emphasis

*italic*   **bold**

_italic_   __bold__
Headers

# Header 1

## Header 2

### Header 3
Lists

Unordered List
* Item 1
* Item 2
    + Item 2a
    + Item 2b
Ordered List
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
Manual Line Breaks

End a line with two or more spaces:
Roses are red,   
Violets are blue.

Blockquotes

A friend once said:

> It's always better to give 
> than to receive.
R Code Blocks

R code will be evaluated and printed
```{r}
summary(cars$dist)
summary(cars$speed)
```
Inline R Code

There were `r nrow(cars)` cars studied
Plain Code Blocks

Plain code blocks are displayed in a fixed-width font but not evaulated
```
This text is displayed verbatim / preformatted
```
Inline Code

We defined the `add` function to 
compute the sum of two numbers.
LaTeX Equations

Inline Equation
$equation$
Display Equation
$$ equation $$
Horizontal Rule / Page Break

Three or more asterisks or dashes:
******

------
Tables

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
-----


Miscellaneous

superscript^2^

~~strikethrough~~










