---
title: "c9w4p1"
author: "Robert Handsfield"
date: "11/10/2015"
output: html_document
---

# S1
Make an interactive widget and publish via rStudio

Lissajous App on shiny.io:
https://r-handsfield.shinyapps.io/lissajous_app

Source code on Github:
https://github.com/r-handsfield/datasciencecoursera/tree/master/9-developingDataProducts/courseProje...

Documentation on RPubs:
http://rpubs.com/r-handsfield/130541

Documentation source on Github:
https://github.com/r-handsfield/datasciencecoursera/tree/master/9-developingDataProducts/courseProject/slidify_documentation


## S 1.1

A shiny app requires two files: `ui.r` and `server.r`.

Run a shiny app with `shiny::runApp(path, ...)`. The `ui.r` and `server.r` files must be in the same directory.

Cancel a shiny app with __Esc__ or __Ctrl+C__.

* ui.r
	+ specifies layout and widget bindings
* server.r 
	+ does the calculations

> OR

---

A shiny app can have one file called `app.r`; this file has the UI and Server stored into variables, that are ultimately called by the `shinyApp()` function at the end of the file. 

Call via `runApp()`??

The `app.r` basic structure looks like this:
```{r eval=FALSE}
library("shiny")

server <- function(input, output) {
	
		# must save result to struct 'output'
  		output$answer <- renderPlot(plot( input * 2 ));
	  }

ui <- fluidPage(
		sidebarPanel(...),
		
		# must call correct field of struct 'output'
    		mainPanel(plotOutput(vector))
	)

# set the 'ui' and 'server' args of shinyApp()
shinyApp(ui = ui, server = server)

```


	
## Basic ui.R

The basic _ui.r_ structure calls `shinyUI()` with 'panel' defining functions as args. Ex. `headerPanel()`, `mainPanel()` etc.

```{r eval=FALSE}
# an example shiny app -- 2015/11/10
# from c9w1: Shiny 1 Introduction to Shiny(8:36)

shinyUI(
	pageWithSidebar(
		headerPanel("Data Science FTW!"),
		sidebarPanel( h3('Sidebar text') ), # as a level 3 html heading
		mainPanel( h3('Main Panel text') ) # as a level 3 html heading
		)
	)
```

### Basic server.r

The basic _server.r_ structure calls the `shinyServer()` function, which in turn takes any number of defined functions as arguments. These argument functions themselves take `input` and `output` as their own arguments.
```{r eval=FALSE}
library(shiny)

shinyServer(
	function(input, output) {
		# function body here
	}
)
```

## next level ui.r

### ui.r html markup

`ui.r` has functions to render html markup tags.

```{r}
library("shiny")
shinyUI(
	pageWithSidebar(
		headerPanel("Illustrating markup"),
		sidebarPanel(
			h1('Sidebar panel'),
			h1('H1 text'),
			h2('H2 Text'),
			h3('H3 Text'),
			h4('H4 Text')
		),
		mainPanel(
			h3('Main Panel text'),
			code('some code'),
			p('some ordinary text')
		)
	)
);

```

### ui.r input bindings

Use one of the `Input()` functions to bind a variable. `...Input()` functions take __html ID, label, value,__ and __value constraints__ as arguments.

```{r eval=FALSE}

shinyUI(pageWithSidebar(
	headerPanel("Illustrating inputs"),
	sidebarPanel(
		
	numericInput('id1', 
		     'Numeric input, labeled id1', 
		     0, min = 0, max = 10, step = 1),
	
	checkboxGroupInput("id2", 
			   "Checkbox", 
			   c("Value 1" = "1","Value 2" = "2","Value 3" = "3")),
	
	dateInput("date", "Date:")
	),
	
	mainPanel(
		h3('Illustrating outputs'),
		h4('You entered'),
		verbatimTextOutput("oid1"), # TextOutput args are defined in server.R
		h4('You entered'),
		verbatimTextOutput("oid2"),
		h4('You entered'),
		verbatimTextOutput("odate")
	)
	)
)
```

### The corresponding server.r file

Note that inputs and outputs referenced by ui.r are named here. Each is a member of the `input` or `output` object.
```{r eval=FALSE}
shinyServer(
	function(input, output) {
		output$oid1 <- renderPrint({input$id1})
		output$oid2 <- renderPrint({input$id2})
		output$odate <- renderPrint({input$date})
	}
)
```

#### Example: A prediction app

Predicts diabetes risk based on blood glucose measurement.

ui.r
```{r eval=FALSE}
shinyUI(
pageWithSidebar(
# Application title
headerPanel("Diabetes prediction"),
sidebarPanel(
numericInput('glucose', 'Glucose mg/dl', 90, min = 50, max = 200, step = 5),
submitButton('Submit')
),
mainPanel(
h3('Results of prediction'),
h4('You entered'),
verbatimTextOutput("inputValue"),
h4('Which resulted in a prediction of '),
verbatimTextOutput("prediction")
)
)
)
```

server.r
```{r eval=FALSE}
diabetesRisk <- function(glucose) glucose / 200
shinyServer(
function(input, output) {
output$inputValue <- renderPrint({input$glucose})
output$prediction <- renderPrint({diabetesRisk(input$glucose)})
}
)
```



# S2 Using Images in Shiny Apps

```{r eval=FALSE}
library("shiny")

# ui.r ########################
shinyUI(pageWithSidebar(
	headerPanel("Example plot"),
	sidebarPanel(
		sliderInput('mu', 'Guess at the mean',value = 70, min = 62, max = 74, step = 0.05,)
	),
	mainPanel(plotOutput('newHist'))
	)
)


# server.r ####################

library(UsingR)
data(galton)

shinyServer(
	function(input, output) {
		output$newHist <- renderPlot({
			hist(galton$child, xlab='child height', col='lightblue', main='Histogram')
			mu <- input$mu
			lines(c(mu, mu), c(0, 200),col="red",lwd=5)
			mse <- mean((galton$child - mu)^2)
			text(63, 150, paste("mu = ", mu))
			text(63, 140, paste("MSE = ", round(mse, 2)))
		})
	}
)
```

# Advanced style control

Instead of doing all the styling in ui.r, you can replace it with an `index.html` (in a new www dir??), which allows all the styling that html/js/css can do. This requires doing all the UI bindings in html or js, probably.

> @TODO find the reference for this

# Distributing a shiny app

An easy way to share with experienced R users is to send/post the dir with all your stuff in it. Then somebody can point R to that dir and call `runApp('./')`.

You can also run your own Shiny server. see [http://www.rstudio.com/shiny/server/](http://www.rstudio.com/shiny/server/) for instructions.

Easier ways to run from a Shiny server

 * [https://www.shinyapps.io/](https://www.shinyapps.io/)
 * [https://sense.io/](https://sense.io/)??


# Advanced Shiny Server files

## Reactive pages

Reactive expressions inside `server.r` execute automatically every time their inputs change.

Basic structure of Shiny server is below. Note when various code blocks are run.
```{r eval=FALSE}

# code before shinyServer(...) runs exactly once, when the server 
# first serves the app/ page

shinyServer(
	function(input, output) { # this is a static function
		
		# code in the static generic function runs every time
		# the app/ page is refreshed
		
		reactive_function_1({ 
			# code inside a reactive (ex renderPlot() ) function runs 
			# every time a widget (button, slider, etc.) invokes the function
		})
		
		reactive_function_2({ })
	}
)
```

### Example 1 Reactive Page

This app increments a number (execute via `runApp(path)`)

> @TODO try runnig with `runApp(path, display.mode='showcase'` to see under-the-hood execution

```{r eval=FALSE}

# ui.R ########################

shinyUI(pageWithSidebar(
	headerPanel("Hello Shiny!"),
	sidebarPanel(
		textInput(inputId="text1", label = "Input Text1"),
		textInput(inputId="text2", label = "Input Text2")
	),
	
	mainPanel(
		p('Output text1'),
		textOutput('text1'),
		p('Output text2'),
		textOutput('text2'),
		p('Output text3'),
		textOutput('text3'),
		p('Outside text'),
		textOutput('text4'),
		p('Inside text, but non-reactive'),
		textOutput('text5')
	)
))

# server.R ####################

# Set x <- 0 before running
# x <- 0

library(shiny)

# this code runs the first time the app is served
x <<- x + 1 # '<<-' keeps the value in the global scope??
y <<- 0

shinyServer(
	function(input, output) {
		y <<- y + 1 # this increments y on every page refresh
		
		# these run every time a text input is changed
		output$text1 <- renderText({input$text1}) 
		output$text2 <- renderText({input$text2})
		output$text3 <- renderText({as.numeric(input$text1)+1})
		output$text4 <- renderText(y)
		output$text5 <- renderText(x)
	}
)

```





