---
title: "c9w2 Lectures"
author: "Robert Handsfield"
date: "11/21/2015"
output: html_document
---
<style>
	div { background-color: #b4b9ff;}
</style>

> NOTE!! Slidify files are .Rmd files. ___Knitting___ a slidify `index.Rmd` ___may___ cause all other .Rmd docs to render as a slideshow rather than web page. Resetting this behavior may require full package removals + rStudio reboot.


# Slidify

The Slidify R packages combines R, knitr, markdown, javascript, and html5 into a single interface for making presentations.

To use Slidify, install the following dependencies:

* `devtools`
* `slidifyLibraries`
* `slidify`

>  <font color='red'> _Slidifiy must be used from inside rStudio!!!_ </font>

```{r message=FALSE, results='hide'}

if (!'devtools' %in% installed.packages()[,'Package']) {install.packages('devtools')}

library(devtools)

# ramnathv::github_install is a vectorised package installer

if (!'slidifyLibraries' %in% installed.packages()[,'Package']) {
	install_github('ramnathv/slidifyLibraries')
}

if (!'slidify' %in% installed.packages()[,'Package']) {
	install_github('ramnathv/slidify')
}
 
library(slidify)
```

> Create a project from the _File_ menu, or with `author("project_name")`.  

> Run a slidify from rStudio with `slidify('index.Rmd')`. 

`author("project_name")` creates:

* d New_Dir/
* d New_Dir/assets/
* d New_Dir/assets/{css, img, js, layouts}
* f New_Dir/index.Rmd

Make new slides by adding markdown `---` (page break) elements.

Add CSS selectors to a slide by labeling the page break with the elements:

`--- .css_class_name #css_id_name`

## The YAML Header

YAML stands for _Yet Another Markdown Language_. YAML code specifies document attributes by using the format

> attribute : value

Index.Rmd has this YAML header, Rmd that has the `knit : slidify::knit2slides` attribute will render as Slidify.
```{r eval=FALSE}
---
title       :   ## title of presentation
subtitle    :   ## subtitle of presentation
author      :   ## author name
job         :   ## author's vocation
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}  ## code highlighter
hitheme     : tomorrow      # {tomorrow, zenburn}                 ## theme for code highlighter
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---
```

Regarding code highlighter themes, the only few that seem to work are:

* tomorrow
	+ light, polychrome, border
* zenburn
	+ dark, polychrome, border
* prettify
	+ big, blank lines, line nums

The following frameworks (slide themes) should work, but only `io2012`, `html5slides`, and `shower` seem to.

* io2012
* shower
* html5slides
* deck.js
* dzslides
* landslide
* Slidy

Other available slidify fields are 
```{r eval=FALSE}
logo        : path_to_image_file
url         : path_to_working_dir
assets      : ../assets # (default value)

```

Also check [http://slidify.org/samples/deck.js/#slide-1]("http://slidify.org/samples/deck.js/#slide-1")

## Publishing Slide Decks

### Publishing to GitHub

Use `slidify::publish_github(repo, username = getOption("github.user"))`

A local Git installation and SSH access to GitHub are required for this.  
The default arg for `username` should work if you have Git/GitHub set up in RStudio.

### Publishing to RPubs

Use `publish_rpubs(title, html_file = "index.html")`

This should be set up and ready to go.


### Publishing to Dropbox

Use `publish_dropbox(dirname)`

This may require username and oath keys??


## Interactive Slides

### Quiz Template

Use the following R markdown, which renders the question, 4 answer choices (_2_ marks the correct choice), and buttons for _Submit_, _Show Hint_, _Show Answer_, and _Clear_:

--- 

## Question 1

What is 1+1?

1. 1
2. _2_
3. 3
4. 4

.hint _Hint text here_

.explanation _Explanatory text here_

---

### Embedding R

Use Rmd just like knitr


# R Studio Presenter

@TODO






