---
title: "c8w4p1"
author: "R. Handsfield"
date: "September 24, 2016"
output: 
  html_document: 
    highlight: pygments
    number_sections: yes
    theme: cerulean
    toc: yes
---

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = 'center', fig.height=3, fig.width=5)
```

```{r library, echo=FALSE, message=FALSE}
library(ggplot2); library(gridExtra)
library(caret)
library(doParallel)
```

# Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


## Download data
```{r download data}
if ("courseProject1" %in% list.files()) {
	setwd("courseProject1")
}

if ( !("pml-training.csv" %in% list.files()) ) {
	urlTrn <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
	download.file(urlTrn, destfile = "pml-training.csv")
}

if ( !("pml-testing.csv" %in% list.files()) ) {
	urlTst <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
	download.file(urlTst, destfile = "pml-testing.csv")
}

# read empty cells and char 'NA' as NA
dfTrn <- read.csv(file = "pml-training.csv", na.strings = c("", "NA"));
dfTst <- read.csv(file = "pml-training.csv", na.strings = c("", "NA"));
```

## Examine data frames
```{r}
names(dfTrn)
dim(dfTrn)
# head(dfTrn, 2L)
# head(dfTst, 2L)
```

### Remove columns with missing data
```{r clean}
keepers <- apply(X = dfTrn, MARGIN = 2, FUN = function(df) !any(is.na(df)) )
dfTrn <- dfTrn[, keepers]

keepers <- apply(X = dfTst, MARGIN = 2, FUN = function(df) !any(is.na(df)) )
dfTst <- dfTst[, keepers]

dim(dfTrn)
names(dfTrn)
```
We have removed 100 columns of NA data. Of the remaining columsn, the first 7 are not important for classification, and column 60 is the outcome.

```{r}
dfTrn <- dfTrn[, 8:60]
dfTst <- dfTst[, 8:60]

```


### Subset

```{r segment}
vecTrn <- createDataPartition(y = dfTrn$classe, p = 0.6, list = FALSE)

val <- dfTrn[-vecTrn,]
trn <- dfTrn[vecTrn, ]
```

```{r complete cases}
trn <- trn[complete.cases(trn), ]
val <- val[complete.cases(val), ]
tst <- dfTst[complete.cases(dfTst), ]
```

## Solution

```{r}
unique(dfTrn$user_name)
unique(dfTrn$classe)
```
There are data for 6 users doing the exercise in 5 different ways (A-E). This should be straightforward - train a classifier that predicts _classe_ as a function of all other variables. 
Try a random forest and gradient boosting model.

```{r single, eval=FALSE, echo=FALSE, cache=TRUE, message=FALSE}
begin <- Sys.time()
m <- train(form = classe~., data = trn, method = "rf")
finish <- Sys.time()

single <- finish-begin
```

Train the models. Enable parallel processing to make this tolerable.
```{r doParallel, message=FALSE}
cl <- makeCluster(spec = 3)
registerDoParallel(cl)

begin <- Sys.time()
modRf <- train(form = classe~., data = trn, method = "rf")
rfTime <- Sys.time() - begin
rfTime

begin <- Sys.time()
modGbm <- train(form = classe~., data = trn, method = "gbm")
gbmTime <- Sys.time() - begin
gbmTime

stopCluster(cl)
```



Make predictions with models
```{r predict, cache=TRUE, message=FALSE}
predRf <- predict(object = modRf, newdata = val)
predGbm <- predict(object = modGbm, newdata = val)
```

Compare models
```{r conf, eval=FALSE}
cmRf  <- confusionMatrix(data = predRf,  reference = val$classe)
cmGbm <- confusionMatrix(data = predGbm, reference = val$classe)

round(  cmRf$overall, 2  ); round(  cmGbm$overall, 2  )
round(  t(cmRf$byClass), 2  ); round(  t(cmGbm$byClass), 2  )
```

The random forest performs slightly better. So, what performance do we get on the test set?
```{r test}
pred <- predict(object = modRf, newdata = tst)
cmTst <- confusionMatrix(data = pred, reference = tst$classe)
```

Overall accuracy of the random forest model is $0.956 \pm .003$

`r cmTst$overall`


### Principle Components

We can see the most important features in the model with

```{r}
modRf$finalModel$importance
```

Some rearranging lets us reorder the columns in the training set by decreasing importance.
```{r eval=FALSE}
components <- data.frame(MeanDecreaseGini <- modRf$finalModel$importance)
components$Sensor <- row.names(components)

components <- data.frame(components[order(components$MeanDecreaseGini, decreasing = TRUE), ])

trnPca <- trn[, components$Sensor]
names(trnPca)
# View(trnPca[, components$Sensor])
```

Now we can plot the training time and accuracy as a function of number of features (1-52)

First, initialize a boatload of vectors to hold optimization metrics.
```{r}

validation_curves <- data.frame(
	numFeatures = 1:52,
	trnBeg =  as.POSIXct( rep(NA, 52), origin="1970-01-01" ),
	trnEnd =  as.POSIXct( rep(NA, 52), origin="1970-01-01" ),
	# trnTime = as.POSIXlt( rep(NA, 52) ),  
	trnTime = as.difftime(tim = rep(0, 52), format = "%M", units = "mins"),

	# training errors
	errTrnA = rep(NA, 52),
	errTrnB = rep(NA, 52),
	errTrnC = rep(NA, 52),
	errTrnD = rep(NA, 52),
	errTrnE = rep(NA, 52),
	
	# validation errors
	errValA = rep(NA, 52),
	errValB = rep(NA, 52),
	errValC = rep(NA, 52),
	errValD = rep(NA, 52),
	errValE = rep(NA, 52)
)
```

```{r num features, cache=TRUE}
cl <- makeCluster(spec = 1)
registerDoParallel(cl)
for (iter in 1:3) {
	print( c("Iteration: ", iter)) 
	names(trn[1:iter])
	
	# train a model using `iter` number of classifiers
	beginTime <- Sys.time()
	modTmp <- train(form = classe~., data = trn[, c(1:iter,53)], method = "rf")
	endTime <- Sys.time()
	
	# log training time
	validation_curves[iter, "trnBeg"] <- beginTime
	validation_curves[iter, "trnEnd"] <- endTime
	# validation_curves[iter, "trnTime"] <- endTime - beginTime
	validation_curves[iter, "trnTime"] <- difftime(endTime, beginTime, units = "min")
	
	print(beginTime)
	print(endTime)
	print(difftime(endTime, beginTime, units="min"))
	print(c("Time = ", validation_curves$trnTime[iter]))
	
	print(modTmp$finalModel$xNames)
	
	# log training error - found in column 6 of the final model confusion matrix
	# the error used here is (1 - Sensitivity)
	validation_curves[iter, 5:9] <- modTmp$finalModel$confusion[,6]
	
	# predict on the validation set
	predTmp <- predict(object = modTmp, newdata = val[, c(1:iter, 53) ])
	
	# calculate Confusion Matrix
	cmTmp <- confusionMatrix(data = predTmp, reference = val$classe)
	
	# log the validation error ( 1- Sensitivity)
	# Sensitivity is in column 1 of confusion_matrix$byClass
	validation_curves[iter, 10:14] <- (1 - cmTmp$byClass[,1])
	
	saveRDS(object = validation_curves, file = "validation_curves.RData")
	
	print("\n\n==============================================\n\n")
}
stopCluster(cl)
```




# Appendix I: Z-Transform

What happens to the models if we z-transform the data sets?
Running `lapply(dfTrn,class)` shows that all the columns are numeric, so we can apply the z-transform to the entire data frame.

The following code would test the models on z-transformed data. Actually running it shows that normalization worsens model performance by $.01 - .03$. No need to run it again because it takes forever.
```{r zTransform, eval=FALSE, cache=TRUE}
trnZ[, 1:52] <- apply(trn[,1:52], 2, scale )
valZ[, 1:52] <- apply(val[,1:52], 2, scale )
tstZ[, 1:52] <- apply(tst[,1:52], 2, scale )

# train the models ##########################
cl <- makeCluster(spec = 3)
registerDoParallel(cl)

begin <- Sys.time()
modRfZ <- train(form = classe~., data = trnZ, method = "rf")
rfTime <- Sys.time() - begin
rfTime

begin <- Sys.time()
modGbmZ <- train(form = classe~., data = trnZ, method = "gbm")
gbmTime <- Sys.time() - begin
gbmTime

stopCluster(cl)

# Make predictions with models ##################################
predRfZ <- predict(object = modRfZ, newdata = valZ)
predGbmZ <- predict(object = modGbmZ, newdata = valZ)

# Compare models ###############################
cmRfZ  <- confusionMatrix(data = predRfZ,  reference = valZ$classe)
cmGbmZ <- confusionMatrix(data = predGbmZ, reference = valZ$classe)

round(  cmRfZ$overall, 2  ); round(  cmGbmZ$overall, 2  )
round(  t(cmRfZ$byClass), 2  ); round(  t(cmGbmZ$byClass), 2  )
```


# Appendix 2 : References

[https://archive.ics.uci.edu/ml/datasets/Wearable+Computing%3A+Classification+of+Body+Postures+and+Movements+%28PUC-Rio%29#](https://archive.ics.uci.edu/ml/datasets/Wearable+Computing%3A+Classification+of+Body+Postures+and+Movements+%28PUC-Rio%29#)

[http://www.mldata.org/repository/data/viewslug/wearable-accelerometers-activity/3/](http://www.mldata.org/repository/data/viewslug/wearable-accelerometers-activity/3/)

[http://www.md2c.nl/how-to-build-a-human-activity-classifier-with-azure-machine-learning/](http://www.md2c.nl/how-to-build-a-human-activity-classifier-with-azure-machine-learning/)

[https://gallery.cortanaintelligence.com/Experiment/Human-Activity-Classifier-2](https://gallery.cortanaintelligence.com/Experiment/Human-Activity-Classifier-2)



[http://nthturn.com/2015/02/22/prediction-using-random-forests-in-r-an-example/](http://nthturn.com/2015/02/22/prediction-using-random-forests-in-r-an-example/)

[https://rstudio-pubs-static.s3.amazonaws.com/19668_2a08e88c36ab4b47876a589bb1d61c37.html](https://rstudio-pubs-static.s3.amazonaws.com/19668_2a08e88c36ab4b47876a589bb1d61c37.html)

[https://yoke2.github.io/PMLCourseProject/pmlreport.html](https://yoke2.github.io/PMLCourseProject/pmlreport.html)

