---
title: "Sample Estimates vs the Central Limit Theorem"
author: "R. Handsfield"
date: "February 19, 2016"
output: 
  pdf_document:
    latex_engine: xelatex
---


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2);
library(gridExtra);

set.seed(15112);
```


# Abstract

In this project, we compare the [exponential distribution](https://en.wikipedia.org/wiki/Exponential_distribution) to the Central Limit Theorem. Simulated exponential distributions of $n = 40$ are sampled from an exponential distribution with rate parameter $\lambda = 0.2$.

# Overview

The exponential distribution, also known as the _Poisson_ distribution, is often
used to measure rates. The probability density and cumulative probability distributions are 

$$
f(x;\lambda) = \left\{\begin{matrix} 
 \lambda e^{- \lambda x} & x \geq 0 \\ 
 0 & x < 0 
\end{matrix}\right.
 ~~~~~~~~~~ ~~~~~~~~~~
F(x;\lambda) = \left\{\begin{matrix} 
 1 - e^{- \lambda x} & x \geq 0 \\ 
 0 & x < 0
\end{matrix}\right.
$$

The parameter _lambda_ is known as the _rate parameter_, and is directly related to the mean and standard deviation.
$$\mu = \frac{1}{\lambda} ~~~~~~~~~~ ~~~~~~~~~~ \sigma = \frac{1}{\lambda}$$


Plots of the exponential distribution look like this
```{r echo=FALSE, fig.align='center', fig.height=3, warning=FALSE}
x <- seq(0,5,.01)
df <- data.frame(x, dexp(x), pexp(x), rexp(x) );
names(df) <- c("x", "exp.dens", "exp.dist", "exp.sim");

h <- ggplot(data=df) + aes(x=x, y=exp.dist) + geom_line();
h <- h + ggtitle("Cumulative Distribution") + ylab("P(X ≤ x)");

g <- ggplot(data=df) + aes(x=x, y=exp.dens) + geom_line();
g <- g + ggtitle("Probability Density") + ylab("P(x)");

grid.arrange(g,h, ncol=2, top="");
```


# Simulations

We create each comparison distribution by sampling 40 values randomly from the exponential distribution.
We simulate a total of 1000 comparisons; the mean of each simulation ($\bar X$) is calculated and appended to the vector `means`.


```{r echo=FALSE, cache=TRUE, message=FALSE}
means <- NULL;  # initialize list of means

for( i in 1:1000 ){
# 	print(i);
	sim <- rexp(40, rate=0.2); # simulate 40 values with lambda = 0.2
	means <- c(means, mean(sim));  # calculate means
}
```

```{r echo=FALSE}
M <- signif(mean(means),3);
V <- signif(var(means),3);
```
\pagebreak

* Average of simulated means: `mean(means)` = `r M`
* Variance of simulated means: `var(means)` = `r V`


According to the Central Limit Theorem, the average of these 1000 means should estimate the true population mean $\mu$. The means should be normally distributed, with a variance proportional to the number of values in each simulation (40).
 

## Sample Mean vs Population Mean

The average of simulated means is the ___Sample Mean:___ $\bar X_{\mu} =$ `r M`. 

The population mean of the exponential distribution is
$$~ \mu = \frac{1}{\lambda} = \frac{1}{0.2} = 5.0$$

`r M` seems close to $5.0$, but how close of an estimate is this? We can figure it out by looking at the distribution of simulated means.

The histogram of simulated means is: 
```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=3}
gm <- ggplot() + aes(x=means) + ggtitle("Sample Means"); 
gm <- gm + geom_histogram(aes(y=..density..), color='black', fill='lightblue');
gm <- gm + geom_vline(xintercept = 5, size=2, color='red');

gm
```

The 1000 simulated means are normally distributed and centered near the true population mean $\mu$ (red line). Because the means are normally distributed, we can use normal statistics to determine how good of an estimte `r M` actually is.

The probability of the sample means randomly centering at `r M` is:
```{r}
1 - 2*(pnorm( abs( mean(means)-5 ), mean=0, sd=sd(means), lower.tail = FALSE) )
```

There is a probability of just a few chances out of 100 that our sample mean estimates a non-exponential population mean, but randomly averages to a value near $5.0$.

## Variance of the Sample Mean

According to the central limit theorem, a sample estimates a statistic, and has variance proportional to the sample size $n$.

$$Var(\bar X) = \left( \frac{\sigma}{\sqrt n}\right)^2$$

For the exponential distribution
$$Var(\bar X) = \left( \frac{1}{\lambda \sqrt n} \right)^{2} = \left( 0.2 \sqrt {40}\right)^{-2} = 0.625$$

The variance of the sample mean is
```{r}
var(means);
```
It appears that the sample variance slightly underestimates the population variance.

## Distribution of the Sample Mean

Overlaying a normal distribution plot on the the histogram of means shows that the distribution of means is approximately normal.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=3}
# add normal distribution plot to histograms
gm <- gm + stat_function(fun=dnorm, args=list(mean=5, sd=sd(means)), size=2)
gm
```

# Conclusion

The simulated distributions yield the following statistic.

* Sample Mean: `r M`
* Variance of Sample Mean: `r V`

Both are in good agreement with the predictions of the central limit theorem.

Distribution of simulated means is approximately normal.

\pagebreak

# Appendix 1: Simulating the Exponential Distribution
```{r eval=FALSE, cache=TRUE, message=FALSE}
means <- NULL;  # initialize list of means

for( i in 1:1000 ){
# 	print(i);
	sim <- rexp(40, rate=0.2); # simulate 40 values with lambda = 0.2
	means <- c(means, mean(sim));  # calculate means
}

mean(means)
var(means)
```


# Appendix 2: Histogram of Simulations
```{r eval=FALSE, warning=FALSE, message=FALSE, fig.height=3}
# plot the histogram
gm <- ggplot() + aes(x=means) + ggtitle("Sample Means"); 
gm <- gm + geom_histogram(aes(y=..density..), color='black', fill='lightblue');
gm <- gm + geom_vline(xintercept = 5, size=2, color='red');

gm
```

```{r, eval=FALSE, warning=FALSE, message=FALSE, fig.height=3}
# add normal distribution plot to histograms
gm <- gm + stat_function(fun=dnorm, args=list(mean=5, sd=sd(means)), size=2)
gm
```