---
title: "Effect of Vitamin C on Tooth Growth"
author: "R. Handsfield"
date: "February 19, 2016"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2);
```


# Abstract

In this project, we analyze the `datasets::ToothGrowth` data set to determine the effect Vitamin C supplement and dosage on the average tooth length of guinea pigs. We use exploratory analysis and hypothesis testing to achieve our results.

# Overview

The `ToothGrowth` dataset reports the effect of Vitamin C on tooth growth in guinea pigs. The data frame contains 3 columns representing, _tooth length_, _supplement type_, and _dose (mg)_.

# Exploratory Analysis
```{r}
library("datasets")
data(ToothGrowth)
```

```{r}
dftg <- ToothGrowth
head(dftg, 3L)
str(dftg)
table(dftg[, c('supp','dose')])
```

```{r eval=FALSE, echo=FALSE}
table(dftg[dftg$supp == 'VC', 'dose'])
dftg[dftg$supp == 'OJ', 'dose']
```
There are two treatments (OJ and VC), each administered at 3 different dosages to independant groups of 10 guinea pigs. 

```{r fig.height=3}
g <- ggplot(dftg) + aes(x=factor(dose), y=len, fill=factor(dose)) + geom_boxplot()
g <- g + facet_grid(.~supp) + xlab("Dose (mg)") + ylab("Tooth Length")
g
```


# Comparing Treatments

* Null hypothesis:  $~~ H_0 : \mu_1 = \mu_2$
* Alternative hypothesis:  $~~H_a : \mu_1 \neq \mu_2$ 
	
```{r echo=FALSE}
# subset individual groups
df_oj <- dftg[dftg$supp=='OJ',]
df_vc <- dftg[dftg$supp=='VC',]

df_half <- dftg[dftg$dose==0.5,]
df_one <- dftg[dftg$dose==1,]
df_two <- dftg[dftg$dose==2,]
```

Summary of Mean Tooth Lengths

| Group | Mean Length |
|-------|-------------|
| OJ | `r round(mean(df_oj$len),3)` |
| VC | `r round(mean(df_vc$len),3)` |
| 0.5 mg | `r round(mean(df_half$len),3)` |
| 1.0 mg | `r round(mean(df_one$len),3)` |
| 2.0 mg | `r round(mean(df_two$len),3)` |

The means of the supplement and dosage groups are all distinct, but are they different enough to suggest that each experimental parameter had an effect?

## T-Testing

Comparing standard errors of each experimental group tells us whether we should use equal variance t-tests.
```{r echo=FALSE}
data.frame(group=c('OJ', 'VC', '0.5 mg', '1 mg', '2 mg'), 
	   std.error= c(sd(df_oj$len)/sqrt(30),
			sd(df_vc$len)/sqrt(30),
			sd(df_half$len)/sqrt(20),
			sd(df_one$len)/sqrt(20),
			sd(df_two$len)/sqrt(20)
			)
	)
```
Neither the treatment nor the dosage groups have similar variances, therefore we use unequal variance t-tests.


```{r}
t <- t.test(len~supp, data=dftg, paired=FALSE, var.equal=FALSE)
data.frame(t$estimate, t$p.value)

t <- t.test(len~dose, data=rbind(df_half, df_one), paired=FALSE, var.equal=FALSE)
data.frame(t$estimate, t$p.value)

t <- t.test(len~dose, data=rbind(df_two, df_one), paired=FALSE, var.equal=FALSE)
data.frame(t$estimate, t$p.value)
```

The means of every group-pair are different, with p-values of 0.06 or better.

# Conclusions
Every experimental parameter caused some change in the tooth growth of a guinea pig population. From t-testing group pairs, we obtain the following results.

| Group | Mean Tooth Length | P-Value |
|-------|-------------------|---------|
| OJ | `r round(mean(df_oj$len),3)` | $0.06$ |
| VC | `r round(mean(df_vc$len),3)` | $0.06$ |
||||
| 0.5 mg | `r round(mean(df_half$len),3)` | $1.27 \times 10^{-7}$ |
| 1.0 mg | `r round(mean(df_one$len),3)` | $1.27 \times 10^{-7}$ |
||||
| 1.0 mg | `r round(mean(df_one$len),3)` | $1.91 \times 10^{-5}$ |
| 2.0 mg | `r round(mean(df_two$len),3)` | $1.91 \times 10^{-5}$ |

To obtain these results, we assume that each guinea pig group is experimentally independant, and has a variance unequal to that of any other group.

\pagebreak

# Appendix 1: Subsetting Groups

```{r eval=FALSE}
# subset individual groups
df_oj <- dftg[dftg$supp=='OJ',]
df_vc <- dftg[dftg$supp=='VC',]

df_half <- dftg[dftg$dose==0.5,]
df_one <- dftg[dftg$dose==1,]
df_two <- dftg[dftg$dose==2,]
```

# Appendix 2: Standard Errors of Groups

```{r}
data.frame(group=c('OJ', 'VC', '0.5 mg', '1 mg', '2 mg'), 
	   std.error= c(sd(df_oj$len)/sqrt(30),
			sd(df_vc$len)/sqrt(30),
			sd(df_half$len)/sqrt(20),
			sd(df_one$len)/sqrt(20),
			sd(df_two$len)/sqrt(20)
			)
	)
```

# Appendix 3: T-Test Results

```{r}
t.test(len~supp, data=dftg, paired=FALSE, var.equal=FALSE)
t.test(len~dose, data=rbind(df_half, df_one), paired=FALSE, var.equal=FALSE)
t.test(len~dose, data=rbind(df_two, df_one), paired=FALSE, var.equal=FALSE)
```
