---
title: "Social Costs of Weather Events: 1950 -- 2011"
author: "Robert Handsfield"
date: "02/02/2015"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
---




# Synopsis
This document available at rpubs.com,  http://rpubs.com/r-handsfield/63104

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  This project analyzes U.S. Government weather data to determine which kinds of weather are most damaging.  

To determine the most damaging weather, this project focused on combined injuries and fatalities, collectively known as *casualties*, and *total monetary damage*, which includes crop and property damage.
Data from 1950--2011, cataloguing 985 kinds of weather events, indicates that tornadoes have caused the most casualties, while floods have caused the most monetary damage.


# Data Processing

## Load Packages
Rquired Packages

* SOAR 
* reshape2
* ggplot2

```{r}
library(package = "ggplot2");
library(package = "SOAR");
library(package = "reshape2");
```



## Get the source data
The dataset, codebooks, and supplementary information may be found at the following links.
The analysis was performed on the NOAA Storm Data Set obtained on **February 2^nd^, 2015**.

* NOAA [Storm Data Set]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")
* Storm Data [Documentation]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf")
* NCDC Storm Events [FAQ]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf")
* Proprietary [Codebook]("https://ire.org/nicar/database-library/databases/storm-events/") from IRE
* [Revisions]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf") to the data set


Download the file
```{r downloadSource, echo=TRUE, cache=TRUE}

fileUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2";

if ( !("Stormdata.csv.bz2" %in% list.files()) ) {
	download.file(fileUrl, destfile = "./Stormdata.csv.bz2", method = "curl" );
}
```  

The file is a .bz2 archive, which can be read into an R data frame
```{r readSource, cache=TRUE}
if ( !("dfStorm" %in% ls() ) ) {
	dfStorm <- read.csv(bzfile("Stormdata.csv.bz2"));	
}
```

As a first analysis, examine data frame column names
```{r examineDfStorm}
names(dfStorm);
```

To answer questions about health and economic impact of weather, focus on the following columns:
```{r makeDfEconHealth, cache=TRUE}
cols <- c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP");
dfEconHealth <- dfStorm[,cols];
```

## Which events are most harmful to **public health**? 
Which EVTYPE values correlate to the highest FATALITIES and INJURIES values?

Look at the most harmful weather *events*.
```{r orderByWorst}
# order descending by injuries, using fatalities as a tie breaker
head( dfEconHealth[with(dfEconHealth, order(-INJURIES, -FATALITIES)),]); 

# order descending by fatalities, using injuries as a tie breaker
head( dfEconHealth[with(dfEconHealth, order(-FATALITIES, -INJURIES)),]);
```

Historically, the most injuries during a single event were caused by a TORNADO.  The most fatalities from a single event were caused by HEAT.

For the next step, sum all fatalities and injuries by event type.
```{r aggInjFatByEvt, cache=TRUE}
dfHealth <- aggregate.data.frame(x = list(dfEconHealth$FATALITIES, dfEconHealth$INJURIES), by = list(dfEconHealth$EVTYPE), FUN = sum);
names(dfHealth) <- c("evtype", "fatalities", "injuries");

# a casualty is an injury or fatality
# sumCasualties = injuries + fatalities
dfHealth$sumCasualties <- dfHealth$fatalities + dfHealth$injuries;

# order descending by injuries
head(dfHealth[with(dfHealth, order(-injuries)),]);

# order descending by fatalities
head(dfHealth[with(dfHealth, order(-fatalities)),]); 

# order descending by total casualties
head(dfHealth[with(dfHealth, order(-sumCasualties)),], 15L); 
```


The top 3 weather events that have **injured** the most people since 1950 are
TORNADO, Thunderstorm Wind, FLOOD; while the top 3 events that have **killed** the most people are
TORNADO, EXCESSIVE HEAT, FLASH FLOOD.  A *casualty* is an injury or fatality, so let's focus on the **total casualties**.

Select arbitrarily, weather types that have caused at least 1000 casualties since 1950.  
A visual inspection of the ordered casualty list shows 14 weather types that meet this criterion.
```{r subsetWorstWeather}
# subset the top 14 most harmful weathers
dfHealthWorst <- head(dfHealth[with(dfHealth, order(-sumCasualties)),], 14L); 

# EVTYPE factor levels are in alphabetical order, so
# re-order the event types by ASCENDING total casualties (for easy plotting later)
dfHealthWorst$evtype = factor(dfHealthWorst$evtype, levels = dfHealthWorst$evtype[order(dfHealthWorst$sumCasualties)], ordered=TRUE);

```

We now have a data frame of the most injurious weather events.  It looks like
```{r headDfHealthWorst}
head(dfHealthWorst, 6L);
```
While not shown above, the default factor order is the same as the total casualties.

Prepare the data frame for plotting
```{r prepCasDfForPlot}

# need to reshape the data frame from wide to long to allow easy plotting
dfTmp <- reshape2::melt(dfHealthWorst[1:3])
names(dfTmp) <- c("evtype", "castype", "casualties");

# re-order casualtyType factor so fatality block appears on top in plot
dfTmp <- rbind(dfTmp[15:28,], dfTmp[1:14,]);

# table is now in long form, and evtype factor levels are still in casualty order

# EVTYPE order should be by total casualties, NOT alphabetical  
str(dfTmp);

# copy data to allow facet plotting with and without TORNADO row
wt <- dfTmp; # series with tornado
wot <- dfTmp[c(2:14, 16:28),]; # series without tornado

# add facet factor 
wt$wT <- as.factor("with TORNADO");
wot$wT <- as.factor("without TORNADO");

dfTmp <- rbind(wt,wot);

dfTmpH <- dfTmp;

# clean up workspace
rm(list = c("wt","wot"));

```

For a plot of the most harmful kinds of weather, see [Plot1](#Plot1)


## Which events are most **expensive**? 
Which EVTYPE values correlate to the highest PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP values?

Examine the structure of the economic impact columns PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
```{r examineDfEcon}

dfEcon <- dfEconHealth[,c(2,5:8)]

str(dfEcon);
```

The column names and data types suggest a numeric value PROPDMG, multiplied by a power of 10 PROPDMGEXP.

$$
\mbox{Cash Property Damage} = PROPDMG \times {PROPDMGEXP} \mbox{ with } {PROPDMGEXP} = 10^n
$$
  and
$$
\mbox{Cash Crop Damage} = CROPDMG \times {CROPDMGEXP} \mbox{ with } {CROPDMGEXP} = 10^n
$$

Unfortunately, the exponent columns contain non-numeric entries:
```{r uniqueFactorsDfEcon}
unique(dfEcon$PROPDMGEXP); unique(dfEcon$CROPDMGEXP);
```

Based on the 
[conclusions]("https://class.coursera.org/repdata-011/forum/thread?thread_id=39#comment-66")
of 
[Eddie Song]("https://class.coursera.org/repdata-011/forum/profile?user_id=10852314"), 
the character codes do represent multipliers,  but the numeric codes are buffer overflows from the preceding columns, and symbols (+,-,?) are errors.  From his analysis, we handle these anomalies by omitting any row that has a non-character exponent code, because it is at least two orders of magnitude smaller than any other damage multiplier.

There are only 341 of these problematic rows, comprising less than 0.1% of the data; omitting them should not affect the results.
```{r omitBadCodes}
numRow <-
length(  grep("[0-8]", dfEcon$PROPDMGEXP )  ) + 
length(  grep("[-?+]", dfEcon$PROPDMGEXP )  ) +
length(  grep("[-?+]", dfEcon$CROPDMGEXP )  ) + 
length(  grep("[0-8]", dfEcon$CROPDMGEXP )  );

numRow;

numRow / nrow(dfEcon);
```

Remove rows with non-alphabetical damage exponent codes
```{r removeBadCodeRows}

dfEcon <- dfEconHealth[,c(2,5:8)]

# order of this regex matters; alternatively, try combos of ("[-]" | "[?]" | "[+]")
badRows <- c(  grep("[0-8+?-]", dfEcon$PROPDMGEXP ), grep("[0-8+?-]", dfEcon$CROPDMGEXP )  ); 

# remove unused factor levels part 1
dfEcon$PROPDMGEXP <- as.character(dfEcon$PROPDMGEXP);
dfEcon$CROPDMGEXP <- as.character(dfEcon$CROPDMGEXP);

# subset
dfEcon <- dfEcon[-badRows,];

# clean up workspace
rm(badRows);

# remove unused factor levels part 2
dfEcon$PROPDMGEXP <- as.factor(dfEcon$PROPDMGEXP);
dfEcon$CROPDMGEXP <- as.factor(dfEcon$CROPDMGEXP);

# check factor levels
levels(dfEcon$PROPDMGEXP);
levels(dfEcon$CROPDMGEXP);
```

Interpret damage exponent character codes as follows:

> Hh ==> 1e2  
> Kk ==> 1e3  
> Mm ==> 1e6  
> Bb ==> 1e9  

For convenience, synonymous EXP character codes should be simplified
```{r simplifyChars}
# inspecting str() & unique() shows that only h,m, m,k must be replaced
str(dfEcon); unique(dfEcon$PROPDMGEXP); unique(dfEcon$CROPDMGEXP);

# convert all character values to upper case

# gsub fails here, try df index:
dfEcon[grep("h", dfEcon$PROPDMGEXP), "PROPDMGEXP"] = "H";
dfEcon[grep("m", dfEcon$PROPDMGEXP), "PROPDMGEXP"] = "M";

dfEcon[grep("k", dfEcon$CROPDMGEXP), "CROPDMGEXP"] = "K";
dfEcon[grep("m", dfEcon$CROPDMGEXP), "CROPDMGEXP"] = "M";
```

To make things convenient, create 2 columns `p.dmg.mult` and `c.dmg.mult` containing values for numeric damage multipliers for property and crops. In R, the form **1e6** is classed as numeric, so when possible, use this form in the new multiplier columns.  

```{r createMultipliers}

# initialize new columns
dfEcon$p.dmg.mult <- rep(NA, nrow(dfEcon)); # property damage multiplier
dfEcon$c.dmg.mult <- rep(NA, nrow(dfEcon)); # crop damage multiplier

# insert the correct multiplier values
dfEcon[grep("B", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e9;  
dfEcon[grep("H", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e2;
dfEcon[grep("K", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e3;
dfEcon[grep("M", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e6;  
dfEcon[which(is.na(dfEcon$p.dmg.mult)), "p.dmg.mult"] = 1e0; # no damage code

dfEcon[grep("B", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e9;
dfEcon[grep("H", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e2;
dfEcon[grep("K", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e3;
dfEcon[grep("M", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e6;
dfEcon[which(is.na(dfEcon$c.dmg.mult)), "c.dmg.mult"] = 1e0; # no damage code

# verify the damage exponent factor levels
str(dfEcon);

```

The economic damage data frame now includes numeric damage multipliers
```{r headDfEcon}
head(dfEcon);
```


Create columns for **cash damage**, the product of **damage** and **multiplier**
```{r createCashDmg}
# initialize cash damage columns
dfEcon$p.dmg.cash <- rep(NA, nrow(dfEcon)); # property damage cash value
dfEcon$c.dmg.cash <- rep(NA, nrow(dfEcon)); # crop damage cash value

dfEcon$p.dmg.cash <- dfEcon$PROPDMG * dfEcon$p.dmg.mult;
dfEcon$c.dmg.cash <- dfEcon$CROPDMG * dfEcon$c.dmg.mult;
```

Aggregate monetary damages by weather type
```{r aggCashByEvt, cache=TRUE}
dfEconWorst <- aggregate.data.frame(x = list(dfEcon$p.dmg.cash, dfEcon$c.dmg.cash), by = list(dfEcon$EVTYPE), FUN = sum);
names(dfEconWorst) <- c("evtype", "prop.damage", "crop.damage");
```

Calculate net damage, the sum of property and crop damage
```{r }
# initialize net damage column
dfEconWorst$net.damage <- rep(NA, nrow(dfEconWorst)); # property damage cash value

dfEconWorst$net.damage <- dfEconWorst$prop.damage + dfEconWorst$crop.damage;

# sort df in descending net.damage order
dfEconWorst <- dfEconWorst[with(dfEconWorst, order(-net.damage)),];

# sort evtype levels in ascending net.damage order
dfEconWorst$evtype = with(dfEconWorst, factor(evtype, levels = evtype[order(net.damage)], ordered=TRUE));

# view descending by property damage
head(dfEconWorst[with(dfEconWorst, order(-prop.damage)),], 10L);

# view ascending by crop damage
head(dfEconWorst[with(dfEconWorst, order(-crop.damage)),], 10L);

# view ascending by net damage
head(dfEconWorst[with(dfEconWorst, order(-net.damage)),], 10L);
```

Prepare the data frame for plotting
```{r prepdfEconForPlot}

dfTmp <- dfEconWorst[1:20,1:3];

# reshape from wide to long
dfTmp <- reshape2::melt(dfTmp);
names(dfTmp) <- c("evtype", "dmg.type", "dmg.amount");

# copy data to allow facet plotting with and without FLOOD row
wf <- dfTmp; # series with flood
wof <- dfTmp[c(2:20, 22:40),]; # series without flood

# add facet factor 
wf$wF <- as.factor("with FLOOD");
wof$wF <- as.factor("without FLOOD");

dfTmp <- rbind(wf,wof);

dfTmpE <- dfTmp;

rm(list = c("wf","wof"));
```

For a plot of the most costliest kinds of weather, see [Plot2](#Plot2)


# Results

### Plot1
<a name="Plot1"></a>
```{r plotCasDf}
p <- ggplot(data=dfTmpH, 
       aes(x=evtype, y=casualties, fill=castype)); # separate bars by castype

p <- p + geom_bar(stat="identity"); # draw bar for each castype

p <- p + facet_grid(wT~., # row for each: w/ TORNADO or w/o TORNADO
	     scales = "free_y"); # uniqe y scale for each facet 

p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 8), # rotate x-axis text
	       strip.text.y = element_text(angle=90)); # rotate facet label text

p <- p + ylab("Casualties") + xlab("Weather Type");

p <- p + ggtitle("Deadliest Weather Events Since 1950");

print(p);
```

When it comes to public health, tornadoes are the most damaging weather event, causing almost 10 times as many casualties as the 2^nd^ most dangerous weather, excessive heat.  Omitting the tornado data magnifies the differences between the remaining types.

### Plot2
<a name="Plot2"></a>
```{r plotCashDmg}

p <- ggplot(data=dfTmpE, aes(x=evtype, y=dmg.amount, fill=dmg.type));

p <- p + geom_bar(stat="identity");

p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 6),  strip.text.y = element_text(angle=90)); # rotate facet label text);

p <- p + facet_grid(wF~., # row for each: w/ FLOOD or w/o FLOOD
	     scales = "free_y"); # uniqe y scale for each facet 

p <- p + ylab("Cost ($)") + xlab("Weather Type");

p <- p + ggtitle("20 Costliest Weather Types Since 1950");

print(p);
```

Property damage makes up most of the economic costs due to weather.  Flooding is most expensive, costing twice as much in damage as the next costliest, hurricane/typhoon.  Omitting flooding magnifies the differences between the remaining types.

This document available on rpubs.com,  http://rpubs.com/r-handsfield/63104



