---
title: "PA2_template.Rmd"
author: "Robert Handsfield"
date: "02/15/2015"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
---


Introduction
----------------------------------------------

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


Load Packages
----------------------------------------------

Rquired Packages

* SOAR 
* reshape2
* ggplot2

```{r}
library(package = "ggplot2");
library(package = "SOAR");
library(package = "reshape2");
```



Get the source data
----------------------------------------------

* NOAA [Storm Data Set]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2");
* Storm Data [Documentation]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf");
* NCDC Storm Events [FAQ]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf");

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

The file is a .bz2 archive, which needs to be extracted.
```{r downloadSource, echo=TRUE, cache=TRUE}

fileUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2";

if ( !("Stormdata.csv.bz2" %in% list.files()) ) {
	download.file(fileUrl, destfile = "./Stormdata.csv.bz2", method = "curl" );
}
```  


```{r readSource, cache=TRUE}
if ( !("dfStorm" %in% ls() ) ) {
	dfStorm <- read.csv(bzfile("Stormdata.csv.bz2"));	
}
```



Questions
-----------------------------------------------------

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

By examining variable names,
```{r examineDfStorm}

names(dfStorm);
```



It seems we should focus on the following columns to answer questions about health and economic impact of weather:
```{r makeDfEconHealth, cache=TRUE}
cols <- c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP");
dfEconHealth <- dfStorm[,cols];
```



# 1. Which events are most harmful? ####################
Which EVTYPE values correlate to the highest FATALITIES and INJURIES values?

As a first examination, look at the most harmful weather *events*.
```{r orderByWorst}
# order descending by injuries, using fatalities as a tie breaker
head( dfEconHealth[with(dfEconHealth, order(-INJURIES, -FATALITIES)),]); 

# order descending by fatalities, using injuries as a tie breaker
head( dfEconHealth[with(dfEconHealth, order(-FATALITIES, -INJURIES)),]);
```

Historically, the most injuries during a single event were caused by a TORNADO.  The most fatalities from a single event were caused by HEAT.

For the next step, sum all fatalities and injuries by event type.
```{r aggInjFatByEvt, cache=TRUE}
dfHealth <- aggregate.data.frame(x = list(dfEconHealth$FATALITIES, dfEconHealth$INJURIES), by = list(dfEconHealth$EVTYPE), FUN = sum);
names(dfHealth) <- c("evtype", "fatalities", "injuries");

# a casualty is an injury or fatality
# sumCasualties = injuries + fatalities
dfHealth$sumCasualties <- dfHealth$fatalities + dfHealth$injuries;

# order descending by injuries
head(dfHealth[with(dfHealth, order(-injuries)),]);

# order descending by fatalities
head(dfHealth[with(dfHealth, order(-fatalities)),]); 

# order descending by total casualties
head(dfHealth[with(dfHealth, order(-sumCasualties)),], 15L); 
```


The top 3 weather events that have **injured** the most people since 1950 are
TORNADO, Thunderstorm Wind, FLOOD; while the top 3 events that have **killed** the most people are
TORNADO, EXCESSIVE HEAT, FLASH FLOOD.  A **casualty** is an injury **or** fatality, so let's focus on the *total casualties**.

Select arbitrarily, weather types that have caused at least 1000 casualties since 1950.  
A visual inspection of the ordered casualty list shows 14 weather types that meet this criterion.
```{r subsetWorstWeather}
# subset the top 14 most harmful weathers
dfHealthWorst <- head(dfHealth[with(dfHealth, order(-sumCasualties)),], 14L); 

# the subset metadata still contains 985 factor levels for 'evtype'
# do the following to remove the unused factor levels
# dfHealthWorst <- data.frame(as.matrix(dfHealthWorst));

# dfHealthWorst$injuries <- as.numeric(dfHealthWorst$injuries);
# dfHealthWorst$fatalities <- as.numeric(dfHealthWorst$fatalities);
# dfHealthWorst$sumCasualties <- as.numeric(dfHealthWorst$sumCasualties);

# there are now only 14 evtype factors, but their levels are in alphabetical order
# re-order the event types by ASCENDING total casualties (for easy plotting later)
dfHealthWorst$evtype = factor(dfHealthWorst$evtype, levels = dfHealthWorst$evtype[order(dfHealthWorst$sumCasualties)], ordered=TRUE);

```

We now have a data frame that looks like
```{r headDfHealthWorst}
head(dfHealthWorst, 6L);
```
While not shown above, the default factor order is the same as the total casualties

Prepare the data frame for plotting
```{r prepCasDfForPlot}
# need to reshape the data frame from wide to long to allow easy plotting

dfTmp <- reshape2::melt(dfHealthWorst[1:3])
names(dfTmp) <- c("evtype", "castype", "casualties");

# re-order casualtyType factor so fatality block appears on top in plot
dfTmp <- rbind(dfTmp[15:28,], dfTmp[1:14,]);

# table is now in long form, and evtype factor levels are still in casualty 
# order, which you can verify by 
str(dfTmp);

# copy data to allow facet plotting with and without TORNADO row

wt <- dfTmp; # series with tornado
wot <- dfTmp[c(2:14, 16:28),]; # series without tornado

# add facet factor 
wt$wT <- as.factor("with TORNADO");
wot$wT <- as.factor("without TORNADO");



dfTmp <- rbind(wt,wot);

rm(list = c("wt","wot"));

```

Combining injuries and fatalities, and arbitrarily choosing only events that have caused at least 1000 casualties since 1950, we see that tornadoes are the most harmful by nearly an order of magnitude.  Plotting again, without including tornado casualties, hilights casualty differences from the subsequent 13 most harmful kinds weather events.



Plot
```{r plotCasDf}



p <- ggplot(data=dfTmp, 
       aes(x=evtype, y=casualties, fill=castype)); # separate bars by castype

p <- p + geom_bar(stat="identity"); # draw bar for each castype

p <- p + facet_grid(wT~., # row for each: w/ TORNADO or w/o TORNADO
	     scales = "free_y"); # uniqe y scale for each facet 

p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5), # rotate x-axis text
	       strip.text.y = element_text(angle=90)); # rotate facet label text

p <- p + ggtitle("Deadliest Weather Events Since 1950");

print(p);
```












# 2. Which events are most expensive? ##################
Which EVTYPE values correlate to the highest PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP values?

As a first examination, examine the structure of the economic impact columns PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
```{r examineDfEcon}

dfEcon <- dfEconHealth[,c(2,5:8)]

str(dfEcon);

# order descending by injuries, using fatalities as a tie breaker
# head( dfEcon[with(dfEcon, order(-(PROPDMG + CROPDMG))),]); 

```

The column names and data types suggest a numeric value PROPDMG, multiplied by a power of 10 PROPDMGEXP.

$$
\mbox{Cash Property Damage} = PROPDMG \times {PROPDMGEXP} \mbox{ with } {PROPDMGEXP} = 10^n
$$
  and
$$
\mbox{Cash Crop Damage} = CROPDMG \times {CROPDMGEXP} \mbox{ with } {CROPDMGEXP} = 10^n
$$

Unfortunately, the exponent columns contain non-numeric entries:
```{r uniqueFactorsDfEcon}
unique(dfEcon$PROPDMGEXP); unique(dfEcon$CROPDMGEXP);
```

Based on the [conclusions]("https://class.coursera.org/repdata-011/forum/thread?thread_id=39#comment-66") of [Eddie Song] ("https://class.coursera.org/repdata-011/forum/profile?user_id=10852314"), the character codes do represent multipliers,  but the numeric codes are buffer overflows from the preceding columns, and symbols (+,-,?) are errors.  From his analysis, we handle these anomalies by omitting any row that has a non-character exponent code, because it is at least two orders of magnitude smaller than any other damage multiplier.

Additionally, there are only 341 of these problematic rows, comprising less than 0.1% of the data.
```{r omitBadCodes}
numRow <-
length(  grep("[0-8]", dfEcon$PROPDMGEXP )  ) + 
length(  grep("[-?+]", dfEcon$PROPDMGEXP )  ) +
length(  grep("[-?+]", dfEcon$CROPDMGEXP )  ) + 
length(  grep("[0-8]", dfEcon$CROPDMGEXP )  );

numRow;

numRow / nrow(dfEcon);
```


Interpret damage exponent character codes as follows:

> Hh ==> 1e2
> Kk ==> 1e3
> Mm ==> 1e6
> Bb ==> 1e9


To make things convenient, add 2 columns containing the values of the damage multipliers for property and crops.  The damage multipliers are calculated according to the formulas above.  
In R, the form **1e6** is classed as numeric, so when possible, use this form in the new multiplier columns.  Also add columns for the **cash damage**, the product of **damage** and **multiplier**.


Remove rows with non-alphabetical damage exponent codes
```{r removeBadCodeRows}

dfEcon <- dfEconHealth[,c(2,5:8)]

# order of this regex matters; alternatively, try combos of ("[-]" | "[?]" | "[+]")
badRows <- c(  grep("[0-8+?-]", dfEcon$PROPDMGEXP ), grep("[0-8+?-]", dfEcon$CROPDMGEXP )  ); 

# remove unused factor levels part 1
dfEcon$PROPDMGEXP <- as.character(dfEcon$PROPDMGEXP);
dfEcon$CROPDMGEXP <- as.character(dfEcon$CROPDMGEXP);

# subset
dfEcon <- dfEcon[-badRows,];

# clean up workspace
rm(badRows);

# remove unused factor levels part 2
dfEcon$PROPDMGEXP <- as.factor(dfEcon$PROPDMGEXP);
dfEcon$CROPDMGEXP <- as.factor(dfEcon$CROPDMGEXP);

# check factor levels
levels(dfEcon$PROPDMGEXP);
levels(dfEcon$CROPDMGEXP);
```



Clean the exponent columns, add the exponent value columns
```{r addMultipliers}
# initialize new columns
dfEcon$p.dmg.mult <- rep(NA, nrow(dfEcon)); # property damage multiplier
dfEcon$c.dmg.mult <- rep(NA, nrow(dfEcon)); # crop damage multiplier

# convert all character values to upper case
# inspecting str() & unique() shows that only h,m, m,k must be replaced
# gsub fails here, try df index:

dfEcon[grep("h", dfEcon$PROPDMGEXP), "PROPDMGEXP"] = "H";
dfEcon[grep("m", dfEcon$PROPDMGEXP), "PROPDMGEXP"] = "M";

dfEcon[grep("k", dfEcon$CROPDMGEXP), "CROPDMGEXP"] = "K";
dfEcon[grep("m", dfEcon$CROPDMGEXP), "CROPDMGEXP"] = "M";


# insert the correct multiplier values
dfEcon[grep("B", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e9;
dfEcon[grep("H", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e2;
dfEcon[grep("K", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e3;
dfEcon[grep("M", dfEcon$PROPDMGEXP), "p.dmg.mult"] = 1e6;
dfEcon[which(is.na(dfEcon$p.dmg.mult)), "p.dmg.mult"] = 1e0; # no damage code

dfEcon[grep("B", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e9;
# dfEcon[grep("H", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e2;
dfEcon[grep("K", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e3;
dfEcon[grep("M", dfEcon$CROPDMGEXP), "c.dmg.mult"] = 1e6;
dfEcon[which(is.na(dfEcon$c.dmg.mult)), "c.dmg.mult"] = 1e0; # no damage code

# verify the damage exponent factor levels
str(dfEcon);

```

We now have a data frame that looks like this
```{r headDfEcon}
head(dfEcon);
```


Add columns for cash damage, and calculate
```{r calcCashDmg}
# initialize cash damage columns
dfEcon$p.dmg.cash <- rep(NA, nrow(dfEcon)); # property damage cash value
dfEcon$c.dmg.cash <- rep(NA, nrow(dfEcon)); # crop damage cash value

dfEcon$p.dmg.cash <- dfEcon$PROPDMG * dfEcon$p.dmg.mult;
dfEcon$c.dmg.cash <- dfEcon$CROPDMG * dfEcon$c.dmg.mult;
```

Aggregate monetary damages by weather type
```{r aggCashByEvt, cache=TRUE}
dfEconWorst <- aggregate.data.frame(x = list(dfEcon$p.dmg.cash, dfEcon$c.dmg.cash), by = list(dfEcon$EVTYPE), FUN = sum);
names(dfEconWorst) <- c("evtype", "prop.damage", "crop.damage");
```

Calculate cash damage
```{r }
dfEconWorst$net.damage <- dfEconWorst$prop.damage + dfEconWorst$crop.damage;

# sort df in descending net.damage order
dfEconWorst <- dfEconWorst[with(dfEconWorst, order(-net.damage)),];

# sort evtype levels in ascending net.damage order
dfEconWorst$evtype = with(dfEconWorst, factor(evtype, levels = evtype[order(net.damage)], ordered=TRUE));

# order descending by property damage
head(dfEconWorst[with(dfEconWorst, order(-prop.damage)),], 10L);

# order ascending by crop damage
head(dfEconWorst[with(dfEconWorst, order(-crop.damage)),], 10L);

# order ascending by net damage
head(dfEconWorst[with(dfEconWorst, order(-net.damage)),], 10L);
```

Prepare the data frame for plotting
```{r prepdfEconForPlot}

dfTmp <- dfEconWorst[1:20,1:3];

# reshape from wide to long
dfTmp <- reshape2::melt(dfTmp);
names(dfTmp) <- c("evtype", "dmg.type", "dmg.amount");

# copy data to allow facet plotting with and without FLOOD row
wf <- dfTmp; # series with flood
wof <- dfTmp[c(2:20, 22:40),]; # series without flood

# add facet factor 
wf$wF <- as.factor("with FLOOD");
wof$wF <- as.factor("without FLOOD");

dfTmp <- rbind(wf,wof);

rm(list = c("wf","wof"));
```

Plot
```{r plotCashDmg}

p <- ggplot(data=dfTmp, aes(x=evtype, y=dmg.amount, fill=dmg.type)); # separate bars by castype

p <- p + geom_bar(stat="identity"); # draw bar for each castype

p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),  strip.text.y = element_text(angle=90)); # rotate facet label text);

p <- p + facet_grid(wF~., # row for each: w/ FLOOD or w/o FLOOD
	     scales = "free_y"); # uniqe y scale for each facet 

p <- p + ylab("Cost ($)") + xlab("Weather Type");

p <- p + ggtitle("20 Costliest Weather Types Since 1950");

print(p);
```

Property damage makes up most of the economic costs due to weather.  Flooding is costliest, costing twice as much in damage as the next costliest, hurricane/typhoon.  Omitting flooding magnifies the differences between the remaining types.



Title
----------------------------------------------
Your document should have a title that briefly summarizes your data analysis

Synopsis
----------------------------------------------
Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

Data Processing
----------------------------------------------
There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.


Results
----------------------------------------------


