---
title: "c9w3l2_Classes_and_Methods"
author: "Robert Handsfield"
date: "01/06/2016"
output: 
  html_document: 
    toc: yes
---

# First

R is interactive AND fully supports object oriented programming. Instead of OOP, it is called the _Classes and Methods System_. OOP in R is fairly different than in other languages, so PAY ATTENTION!!!

Because R inherits from the S Language, R has both S3 and S4 classes and methods

* S3 classes/methods
	+ informal, klugey
	+ can be called directly - but don't ever do it
	+ most base R functions are S3 style
	+ sometimes called _old-style_ methods by old-timers
* S4 classes/methods
	+ formal and rigorous
	+ cannot be called directly
	+ sometimes called _new-style_ by kool kids
	+ implementation code in the R `methods` package
	
Along with classes, objects, and methods, R has _generic functions_, which are important for some reason. Generic functions figure what data type they are receiving, then call appropriate methods on that data; they DON'T DO ANYTHING ELSE.

To be a smart R programmer, read the documentation from 

* `?Classes`
* `?Methods`
* `?setClass`
* `?setMethod`
* `?setGeneric`


## Second

Classes

* Classes defined with `methods::setClass()`
* Instantiated by `methods::new()`

Methods

* Class methods defined with `methods::setMethod`

Generic Function
* defined with `methods::setGeneric()`


## Third

The class of all objects can be determined with the `class()` function. The most basic classes, like string, int, etc, are called _atomic_ classes. More complicated data objects (duh) can have their class types returned in the same manner.

```{r}
# atomic classes
class(16);
class("a");
class(5.3);

# data classes
x <- rnorm(100);
y <- rnorm(100);
fit <- lm(y ~ x); ## Linear regression model
class(fit);
```

## Fourth
Basic information about methods can be printed by typing the method name.
```{r}
mean
```


Methods can be investigated more deeply with `utils::methods()`. Calling `methods` on the generic function `base::mean()` shows that `mean()` can call five different averaging methods, one each for: dates, numbers, time differences, POSIXct, POSIXlt. It all depends on the class of the input data, and `base::mean()` figures this out for us.
```{r}
# an S3 generic function
methods("mean");
```

Loading new packages can add methods called by `mean()`.


## Fifth

The S4 generic function `methods::show()` (S4 equivalent of print) can also call many different object methods.

```{r}
methods("show")

# or 
# showMethods("show")
```

## How Generic Functions Work

The mechanism for methods/ generic functions is

0. The first argument of a generic function is an object of a particular class (there may be other
arguments)
1. The generic function checks the class of the object.
2. A search is done to see if there is an appropriate method for that class.
3. If there exists a method for that class, then that method is called on the object and we’re done.
4. If a method for that class does not exist, a search is done to see if there is a default method for
the generic. If a default exists, then the default method is called.
5. If a default method doesn’t exist, then an error is thrown.
	
## Examining Method Code

Most method code is hidden, but you can view it like this

* S3: `getS3method(<name of generic>, <class>)`
	+ `utils::getS3ethod()`
* S4: `getMethod(<name of generic>, <signature>)`
	+ `methods::getMethod()`


## Creating S4 Classes

S4 classes are defined with `methods::setClass()`. At minimum, the class needs a name. 

`setClass` example
```{r eval=FALSE}
setClass(Class, representation, prototype, contains=character(), validity, 
	 access, where, version, sealed, package, S3methods = FALSE, slots)
```

The arguments are, in order of priority

1. Class = name of the class
2. slots = class variables in format `list(variable="type", etc)`
3. contains = superclasses in format `list("superclass")` 
	+ superclasses must be in same type-order as slot elements


Class variables are called _slots_ and are statically typed. Slots are defined via the `slots` argument of `setClass()` like this
```{r eval=FALSE}
# create list of slot='type' with list() or vector()
classVars = list(x="numeric", y="integer", m="lm", b="logical")

# create a list of parent classes (optional)
superClass = list("numeric", "integer", "lm")

setClass(Class = "myClass", slots = classVars, contains = superClass ...)
```



* Define new class with `methods::setClass()`
* Define slots (class variables) with the `slots` argument of `setClass()`
	+ or use `methods::representation()` (almost deprecated)
	+ slots are accessible with the `@` operator
* Define methods with `methods::setMethod()`
* View class structure with `methods::showClass()`
	
### Example - A Polygon Class

```{r eval=FALSE}
library(methods)

setClass("polygon", slots = c(x="numeric", y = "numeric") )
```

## Creating S4 Methods

S4 methods are created with he `setMethod()` function 
```{r eval=FALSE}
setMethod(f, signature=character(), definition, where = topenv(parent.frame()), 
          valueClass = NULL, sealed = FALSE)
```

1. `f` = A generic function or the character-string name of the function
2. `signature` = a list of objects the function can accept
3. `definition` = the function definition, standard syntax

`setMethod()` requires both a signature __AND__ an associated generic function (primitive base function like `mean` or `print`). The signature is a character vector indicating the types of objects that are accepted by the method. For example, `mean()` accepts numerics, integers, dates, date ranges, etc.

Working from the polygon class in the previous example, add a plotting method that takes a polygon object as input.

```{r}
library(methods)

setClass("polygon", slots = c(x="numeric", y = "numeric"))

setMethod("plot", "polygon",
	  function(x, y, ...) {
	  	plot(x@x, x@y, type="n", ...);
	  	xp <- c(x@x, x@x[1]);
	  	yp <- c(x@y, x@y[1]);
	  	lines(xp, yp);
	  }
	  )
```

We've created an additional method called "plot". Now, R adds our method to the base package generic function for `plot`. We can see the addition with 
```{r}
library(methods)
showMethods("plot")
```

The signature for our class `polygon` is listed. The other signature is for the default plot method.

Let's try it out
```{r}
# new class instance
p <- new("polygon", x = c(1,2,3,4), y = c(1,2,3,1))

plot(p)
```








	
	
	
	
	
	
	
	
	
	
	
	