---
title: "c6w1h1"
author: "Robert Handsfield"
date: "04/16/2015"
output:
  html_document:
    toc: yes
---

# Question 1

>Consider influenza epidemics for two parent heterosexual families. Suppose that the probability is 15% that at least one of the parents has contracted the disease. The probability that the father has contracted influenza is 10% while that the mother contracted the disease is 9%. What is the probability that both contracted influenza expressed as a whole number percentage?

a) 15%
b) 10%
c) 9%
d) 4%

[Watch video solution](https://www.youtube.com/watch?v=CvnmoCuIN08&index=1&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L)

## Solution:
$P(M \cap F) = ?$

Need 

$$P(A \cup B) = P(A) + P(B) - P(A \cap B)$$
$$\rightarrow P(M \cup F) = P(M) + P(F) - P(M \cap F)$$

$P(M \cup F) = 15\%$

$P(M) = 9\%$

$P(F) = 10\%$

<br />

$P(M \cap F) = P(M) + P(F) - P(M \cup F)$
$P(M \cap F) = .10 + .09 - .15$

```{r}
.10 + .09 - .15
```

__Answer:__ $P(M \cap F) = 4\%$

<br />

# Question 2

>A random variable, $X$, is uniform, a box from $0$ to $1$ of height $1$. (So that its density is $f(x)=1$ for $0≤x≤1$) What is its median expressed to two decimal places?

a) 1.00
b) 0.75
c) 0.50
d) 0.25

[Watch video solution](https://www.youtube.com/watch?v=UXcarD-1xAM&index=2&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L)

## Solution
```{r echo=FALSE, fig.width=4.2, fig.height=3, fig.align='center'}
library("ggplot2")
ggplot() + aes(x=c(-.5,0,0,1,1,1.5), y=c(0,0,1,1,0,0)) + geom_line() + xlab("") + ylab("") +
	   geom_rect(xmin=0, xmax=1, ymin=0, ymax=1, fill="#FF000006") + 
	   geom_hline(yintercept=0.5, color="blue")
```

The density is uniform, and the median is the value at which _50% of the total density lies below._ The blue line at $y=0.50$ is the median.

__Answer:__ The median is $0.50$

<br />

# Question 3

>You are playing a game with a friend where you flip a coin and if it comes up heads you give her $X$ dollars and if it comes up tails she gives you $Y$ dollars. The odds that the coin is heads is $d$. What are your expected earnings?

a) $-X{d \over 1+d} + Y{1 \over 1+d}$
b)  $X{d \over 1+d} + Y{1 \over 1+d}$
c)  $X{d \over 1+d} - Y{1 \over 1+d}$
d) $-X{d \over 1+d} - Y{1 \over 1+d}$

[Watch video solution](https://www.youtube.com/watch?v=5J88Zq0q81o&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L&index=3)

## Solution

* odds of heads = $d$
* heads means $-X$ dollars
* tails means $+Y$ dollars


Win     | I win ($p$)   | You win ($1-p$) 
------- | ------------- | ----------------
$P(w)$	| $p$		| $1-p$

	
Over time, I expect $[Earnings] = Y(1-p) - Xp$	

__Odds $(d)$ are defined as__ $Odds = {p \over 1-p} \therefore  p = {d \over 1 + d}$

Substitute for $p$:  

$$[Earnings] = Y(1 - {d \over 1 + d} - X{d \over 1 + d})$$

$$[Earnings] = Y({1+d - d} \over 1+d) -  X{d \over 1 + d})$$

__Answer:__ $$[Earnings] = Y(1 \over 1+d) -  X{d \over 1 + d})$$
	


# Question 4

>A random variable takes the value -4 with probability .2 and 1 with probability .8. What is the variance of this random variable?

a) 0
b) 4
c) 8
d) 16

[Watch video solution](https://www.youtube.com/watch?v=Em-xJeQO1rc&index=4&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L)

## Solution

The variance is defined in terms of expectated values: $Var(X) = E[X^2] - E[X]^2$

The expected value is the sum of possible values and their probabilities:

$x$	| $p(x)$  | $x^2$ | $p(x^2)$
------- | ------- | ----- | ---------
-4	| 0.2	  | 16	  | 0.2
1	| 0.8	  | 1	  | 0.8

$E[X] = -4 * 0.2 + 1 * 0.8 = 0 \therefore E[X]^2 =0$

$E[X^2] = 16 * 0.2 + 1 * 0.8 = 4.0$

$\sigma^2 = E[X^2] - E[X]^2 = 4.0 - 0 = 4$

```{r}
# var = E[X^2] - E[X]^2
16 * .2 + 1 * .8 - (-4*.2 +1*.8)
```

__Answer:__ Variance = 4

<br />


# Question 5

>If $\bar{X}$ and $\bar{Y}$ are comprised of n iid random variables arising from distributions having means $μ_x$ and $μ_y$, respectively and common variance $σ^2$ what is the variance $\bar{X} - \bar{Y}$? 

a) $0$
b) $2\sigma^2 / n$
c) $\mu_x - \mu_y$
d) $2\sigma^2$ 

[Watch video solution](https://www.youtube.com/watch?v=7zJhPzX6jns&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L&index=5)

## Solution

```{r echo=FALSE, fig.width=4.2, fig.height=3, fig.align='center'}
set.seed(12343);
x <- seq(-10,10,length=1000);
pop <- dnorm(x,mean=0, sd=3);
samp <- dnorm(x, mean=0, sd=1);

library("ggplot2")
ggplot() + geom_line(aes(x=x, y=pop)) + geom_line(aes(x=x, y=samp), color="red");
```
The population distribution in black. A random sample (red) taken from the population will always be narrower and maybe taller. 

The variance of the population is $\sigma^2$ and the variance of the sample mean is $Var(\bar{x}) = {\sigma^2 \over n}$

Since $\bar{X}$ and $\bar{Y}$ are independent, 

$$Var(\bar{X} - \bar{Y}) \rightarrow Var(\bar{X}) - Var(\bar{Y})$$

But, variances __always__ add, so 

$$Var(\bar{X}) + Var(\bar{Y}) = {\sigma^2 \over n} + {\sigma^2 \over n} = {2\sigma^2 \over n}$$

__Answer:__ $Var(\bar{X} - \bar{Y}) = {2\sigma^2 \over n}$

# Question 6

>Let $X$ be a random variable having standard deviation $σ$. What can be said about $X/σ$?


a) Nothing
b) It must have variance 1
c) It must have mean 0
d) It must have variance 0

[Watch video solution](https://www.youtube.com/watch?v=0WUj18_BUPA&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L&index=6)

## Solution

From the statement, $Var(X) = \sigma^2$

We know that $E \left[ {X \over \sigma} \right] = {E[X] \over \sigma}$

What about the variance of $X$? From the identity $Var(aX) = a^2Var(X)$, we get

$Var {X \over \sigma} = {1 \over \sigma^2}Var(X) = {\sigma^2 \over \sigma^2} = 1$ 

__Answer:__ $Var({X \over \sigma}) = 1$



# Question 7

>If a continuous density that never touches the horizontal axis is symmetric about zero, can we say that its associated median is zero?


a) Yes
b) No
c) It can not be determined given the information given


[Watch video solution](https://www.youtube.com/watch?v=sn48CGH_TXI&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L&index=7)

## Solution

__Answer:__ Yes

# Question 8

>Consider the following pmf given in R
```{r}

p <- c(.1, .2, .3, .4)
x <- 2 : 5

```
>What is the variance expressed to 1 decimal place?

a) 1.0
b) 4.0
c) 6.0 
d) 17.0

[Watch video solution](https://www.youtube.com/watch?v=sn48CGH_TXI&list=PLpl-gQkQivXhHOcVeU3bSJg78zaDYbP9L&index=7)

## Solution

$Var(X) = E[X^2] - E[X]^2$

```{r}

p <- c(.1, .2, .3, .4)
x <- 2 : 5
sum(x*x*p) - sum(p*x)^2
```

__Answer:__ $Var(X) = 1.0$

# Question 9


a)
b)
c)
d)

[Watch video solution]

## Solution


# Question 10


a)
b)
c)
d)

[Watch video solution]

## Solution