---
title: "c6w1q1"
author: "Robert Handsfield"
date: "02/09/2015"
output: html_document
---
Question 1
--------------------------------------------
Consider influenza epidemics for two parent heterosexual families. Suppose that the probability is 17% that at least one of the parents has contracted the disease. The probability that the father has contracted influenza is 12% while the probability that both the mother and father have contracted the disease is 6%. What is the probability that the mother has contracted influenza?

(Hints look at lecture 2 around 5:30 and homework question on page 3/10). 

$$ P(M \cup F) = 0.17$$
$$ P(F) = 0.12$$
$$ P(M \cap F) = 0.06$$
$$P(M) = ??$$

Need
$$P(M \cup F) = P(M) + P(F) - P(M \cap F)$$
$$\therefore P(M) = P(M \cup F) + P(M \cap F) - P(F)$$
$$P(M) = 0.17 + 0.06 - 0.12 = 0.11$$

__Answer:__ 0.11

```{r}
# P(M or F) = .17
# P(F) = .12
# P(M and F) = .06

.17 + .06 - .12
```

Question 2
---------------------------------------------------
A random variable, $X$ is uniform, a box from 0 to 1 of height 1. (So that its density is $f(x)=1 \mbox{ for } 0≤x≤1$.) What is its 75th percentile?

(Hints, look at lecture 2 around 21:30 and homework 1 page 4/10. Also, look up the help function for the `qunif` command in R.)

<br />

If (0 < x < 1) and y(x) = 1, at what value of x is 75% of the box area to the left?

Formally, find the random input for the CDF equaling 0.75: $F(x) = 0.75, x = ?$

Also, any region of the uniform distribution will be a rectangle: $A = b \cdot h$

$$F(x) = P(X \leq x) = Base \times 1 = x = 0.75$$

```{r}

qunif(.75, 0 , 1);

```


__Answer:__ It's a fucking box, x = 0.75



Question 3  
--------------------------------------------------
You are playing a game with a friend where you flip a coin and if it comes up heads you give her X dollars and if it comes up tails she gives you Y dollars. The probability that the coin is heads is p (some number between 0 and 1.) What has to be true about X and Y to make so that both of your expected total earnings is 0. The game would then be called “fair”. 

(Hints, look at Lecture 4 from 0 to 6:50 and Homework 1 page 5/10. Also, for further reading on fair games and gambling, start with the [Dutch Book problem](http://en.wikipedia.org/wiki/Dutch_book) ). 

See c6w1q1.svg

__Answer:__ ${p \over {1-p}} = {Y \over X}$  






Question 4   
---------------------------------------------------

A density that looks like a normal density (but may or may not be exactly normal) is exactly symmetric about zero. (Symmetric means if you flip it around zero it looks the same.) What is its median?

(Hints, look at quantiles from Lecture 2 around 21:30 and the problem on page 9/10 from Homework 1.)

<br />

This seems like the median is at zero, but it's only true if the CDF never touches the horizontal axis.
So the "may or may not be exactly normal" line is important, and probably means the answer is
~ can't conclude anything

__Answer:__ As long as the distribution is symmetric, the median is zero.





Question 5
------------------------------------------------------
Consider the following PMF shown below in R
```{r }
x <- 1:4
p <- x/sum(x)
temp <- rbind(x, p)
rownames(temp) <- c("X", "Prob")
temp
```
```{r echo=FALSE}
##      [,1] [,2] [,3] [,4]
## X     1.0  2.0  3.0  4.0
## Prob  0.1  0.2  0.3  0.4
```
What is the mean?

(Hint, watch Lecture 4 on expectations of PMFs and look at Homework 1 problem on page 10/10 for a similar problem calculating the variance.)

$$E[X] = \sum_x xp(x)$$

```{r}
sum(x*p);
```

__Answer:__ 3 



Question 6
---------------------------------------------
A web site (www.medicine.ox.ac.uk/bandolier/band64/b64-7.html) for home pregnancy tests cites the following: “When the subjects using the test were women who collected and tested their own samples, the overall sensitivity was 75%. Specificity was also low, in the range 52% to 75%.” Assume the lower value for the specificity. Suppose a subject has a positive test and that 30% of women taking pregnancy tests are actually pregnant. __What number is closest to the probability of pregnancy given the positive test?__


(Hints, watch Lecture 3 at around 7 minutes for a similar example. Also, there's a lot of Bayes' rule problems and descriptions out there, for example here's one for HIV testing. Note, discussions of Bayes' rule can get pretty heady. So if it's new to you, stick to basic treatments of the problem. Also see Homework 2 question on page 5/12.) 


 * Sensitivity = .75
 * Specificity = .52 
 * P(D) = .3
 	+ aka _Prevalence_
 * Test = +
 * PPD = ?
 
 $$PPD = P(D|+) = { {Sens \times Prev}  \over  {Sens \times Prev + (1 - Spec) \times (1 - Prev)} }$$
 
 $$PPD = { {.75 \times .3}  \over { .75 \times .3 + (1 - .52) \times (1 - .3) } \approx 0.40 }$$
 
```{r}
.75 * .3 / ( .75 * .3 + (1 - .52) * (1 - .3))
```

__Answer:__  40%





