---
title: "c6w2q1"
author: "Robert Handsfield"
date: "10/18/2015"
output: html_document
---


# c6w2q1

## Question 1
What is the variance of the distribution of the average an IID draw of n observations from a population with mean μ and variance σ^2.

a) $2 \sigma / \sqrt{n}$
b) $\sigma^2$  
__c) $\sigma^2 \over n$__
d) $\sigma / n$

### Solution

The easy relationship to remember is the standard deviation.

The sample standard deviation $$SD_{\bar X} = \sigma / \sqrt{n}$$ 

The variance is just the square of the standard deviation, so

$$Var_{\bar X} = SD_{\bar X}^2 = (\sigma / \sqrt{n}) = \sigma^2 / n$$

#### Answer: $\sigma^2 \over n$

---

## Question 2

Suppose that diastolic blood pressures (DBPs) for men aged 35-44 are normally distributed with a mean of 80 (mm Hg) and a standard deviation of 10. About what is the __probability__ that a random 35-44 year old has a __DBP less than 70__?

__a) 16%__  
b) 8%  
c) 22%  
d) 32%  

### Solution

* $\mu = 80$
* $\sigma = 10$
* $X = 70$

Because the normal distribution has been thoroughly calculated, Z-transform $X$ to $Z ~ N(0,1)$ to get the quantile. 

$$P(X < 70) = P(Z < \frac{X - \mu}{\sigma}$$
$$P(X < 70) = P(Z < \frac{70 - 80}{10}$$
$$P(Z < -1) = 0.159$$

Or, in R
```{r}
# pnorm does value --> quantile
# qnorm does quantile --> SD value
pnorm(70, mean = 80, sd = 10, lower.tail = TRUE)
```

#### Answer: 0.159

---


## Question 3

Brain volume for adult women is normally distributed with a mean of about 1,100 cc for women with a standard deviation of 75 cc. What brain volume represents the 95th percentile?

a) approximately 1247  
__b) approximately 1223__  
c) approximately 977  
d) approximately 1175  

### Solution

* $\mu = 1100$
* $\sigma = 75$

The $95^{th} percentile of the Normal Distribution lies at $z = \pm 1.645$ ($z$ is termed $K$ in the later section on confidence intervals)

(confirm with)
```{r}
qnorm(0.95)
```

Inverse Z-transform this multiplier by 
$$z = \frac{x - \mu}{\sigma} \to x = \mu + z\sigma$$ 
$$x = 1100 + 1.645 * 75$$

```{r}
1100 + 1.645 * 75
```


#### Answer: 1223

---

## Question 4

Refer to the previous question. Brain volume for adult women is about 1,100 cc for women with a standard deviation of 75 cc. Consider the sample mean of 100 random adult women from this population. What is the 95th percentile of the distribution of that sample mean?

a) approximately 1088 cc  
__b) approximately 1112 cc__  
c) approximately 1110 cc  
d) approximately 1115 cc  

### Solution

By the __Central Limit Theorem__: 

> If one takes random samples of size $n$ from a population of mean $\mu$ and standard deviation $\sigma$, then, as $n$ gets large, the _sample mean_ $\bar X$ approaches the Normal Distribution with mean $\mu$ and __standard deviation $\sigma / \sqrt{n}$__.

therefore

* $\mu = 1100$
* $n = 100$
* $\sigma = 75/ \sqrt{100} = 7.5$

Perform the inverse Z-transform again (Q3)
$$z = \frac{x - \mu}{\sigma} ~ \therefore ~ x = \mu + z\sigma$$ 
$$x = 1100 + 1.645 * 7.5$$

```{r}
1100 + 1.645 * 7.5
```

#### Answer: 1112.338

---


## Question 5

You flip a fair coin 5 times, about what's the probability of getting 4 or 5 heads?

__a) 19%__  
b) 12%  
c) 6%  
d) 3%  

Solution:

* $n = 5$ 
* $X = {4,5}$

From

$$P(X=x) = \binom{n}{x}{p^x}(1-p)^{n-x}$$
$$P(X = {4,5}) = P(4) + P(5)$$

```{r}
# use pbinom *** q > 3 = {4,5}
pbinom(q = 3, size = 4, prob = .5, lower.tail = FALSE)
```

__Answer: $0.1875$__

---

## Question 6

The respiratory disturbance index (RDI), a measure of sleep disturbance, for a specific population has a mean of 15 (sleep events per hour) and a standard deviation of 10. They are not normally distributed. Give your best estimate of the probability that a sample mean RDI of 100 people is between 14 and 16 events per hour?

a) __68%__
b) 95%
c) 47.5%
d) 34%

### Solution

Regardless of the original population's distribution, random samples _always_ approximate the Normal. So, use the standard confidence interval equations for the Normal distribution.

* $\mu = 15$
* $\sigma = 10$
* $n = 100$
* $X = {14, 16}$

---

The standard deviation of the sample is

* $SD_{X} = \frac {\sigma}{\sqrt{n}} = 1$

---

From the definition of confidence interval
$$CI({X}, n) = {\mu} \pm K\sigma / \sqrt n$$

We can find the $K$ multiplier that yields a given confidence interval
$$K = \frac{[CI(X, n) - \mu] \sqrt{n}}{\sigma}$$

(Because this is a _sample_ from a population, we should technically be writing $SD_{X}$ instead of $\sigma$. But, screw it.)

---

$$K = \frac{(\{16, 14\} - 15)*10}{10} = \pm 1$$

The given confidence interval for this sample $\{16, 14\}$ lies within 1 standard deviation $SD_{X}$ of the sample's mean. From the table of Standard Normal values, $Z_N(0,1) = 1 \sim X_N(0,1) = 0.841 ~$ and $~ Z_N(0,1) = -1 \sim X_N(0,1) = 0.159$ 

---

In this particular case, the $X_N(0,1)$ is the percentage of $X$ values _below_ $F(Z_N(0,1) \vert_z)$, so we just take the difference 

$$\sum{X_N(0,1) \vert_{F(Z)}} = F(1) - F(-1) = 0.841 - 0.159 = 0.682$$


#### Answer: $0.682$

---

## Question 7

Consider a standard uniform density. The mean for this density is .5 and the variance is ${1 / 12} = (0.0833)$. You sample 1,000 observations from this distribution and take the sample mean, what value would you expect it to be near?

a) 0.10
b) 0.25
c) 0.75  
__d) 0.5__

### Solution

The uniform distribution is a boxcar function given by $f(x) = \frac{1}{b-a}$. $A$ is the _location parameter_, giving the displacement of the lower bound from 0. $(B-A)$ is the _scale parameter_, giving the doman of the box. The area under the curve is always $1$, meaning that when the domain increases, the function value decreases proportionally. Consequently, the variance is a function of the domain

__The Uniform Distribution__

* Function $f(x) = \frac{1}{b-a}$
* Mean $\mu = \tfrac{1}{2}(a+b)$
* Median $m = \tfrac{1}{2}(a+b)$
* Variance $\sigma^{2} = \tfrac{1}{12}(b-a)^2$

The _standard uniform distribution_ occurs when $a=0$ and $b=1$. In this case the function and domain are $1$, the median and mode are $1/2$, and the variance is $1/12$.

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=4}
x <- c(-.2,-.1,0,0,1,1,1.1,1.2); y <- c(0,0,0,1,1,0,0,0);

plot(x,y, type='l', xlab="a=0, b=1", ylab="f(x)=1", col="blue", main="The Standard Uniform Distribution")
```


#### Answer:
__0.5__

--- 

## Question 8

The number of people showing up at a bus stop is assumed to be Poisson with a mean of 5 people per hour. You watch the bus stop for 3 hours. About what's the probability of viewing 10 or fewer people?

a) 0.03  
b) 0.06  
c) 0.08  
__d) 0.12__  

### Solution

When modeling rates, $\lambda = rate \times time$ $\therefore \lambda = 5 * 3$

(model the function here)

```{r}
ppois(10, lambda = 5*3)
```

#### Answer: 

$0.118 \to$ __$0.12$__

