---
title: "c7w1q1"
author: "Robert Handsfield"
date: "04/12/2015"
output: html_document
---



# Question 1 

Consider the data set given below
```{r}
x <- c(0.18, -1.54, 0.42, 0.95)
```

And weights given by
```{r}
w <- c(2, 1, 3, 1)
```

Give the value of $\mu$ that minimizes the least squares equation $\sum_{i=1}^n{w_i(x_i-\mu)^2}$

a) __0.1471__
b) 0.300
c) 1.077
d) 0.0025

## Solution
Use

```{r}
# finishes with errors
lm(x~x, weights=w);

#Or, plug each answer into sum(w*(x-u)^2) 
sum(w*(x-.1471)^2); 
sum(w*(x-.3)^2); 
sum(w*(x-1.077)^2); 
sum(w*(x-.0025)^2);
```

__Answer 1:__ 0.1471

# Question 2 

Consider the following data set

```{r}
x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42)
y <- c(1.39, 0.72, 1.55, 0.48, 1.19, -1.59, 1.23, -0.65, 1.49, 0.05)
```


Fit the regression through the origin and get the __slope__ treating y as the outcome and x as the regressor. (Hint, do not center the data since we want regression through the origin, not through the means of the data.)

__Slope = ?__

a) 0.59915
b) -0.04462
c) -1.713
d) __0.8263__

## Solution

Use $\hat{\beta_1} = { {\sum_{i=1}^{n}Y_i X_i} \over {\sum_{i=1}^{n}X_i^2} }$
```{r}
sum(x*y)/sum(x^2);
```


__Answer 2:__ 0.8263




# Question 3 

Do `data(mtcars)` from the datasets package and fit the regression model with mpg as the outcome and weight as the predictor. Give the __slope coefficient__.

a) __-5.344__
b) -9.559
c) 0.5591
d) 30.2851

## Solution
```{r}
data(mtcars);

# X = wt, Y = mpg; form is lm(y~x, ...)
lm(formula = mpg ~ wt, data=mtcars);
```

__Answer 3:__ -5.344



# Question 4

Consider data with an outcome (Y) and a predictor (X). The standard deviation of the predictor is one half that of the outcome. The correlation between the two variables is .5. What value would the __slope coefficient__ be for the regression model with Y as the outcome and X as the predictor?

a) 0.25  
b) __1__  
c) 3  
d) 4

## Solution
$Sd(X) = {Sd(Y) \over 2}$

$Cor(Y,X) = 0.5$

$\hat{\beta_1} = ?$

Use   
$\hat{\beta_1} = Cor(Y,X) \cdot { {Sd(Y)} \over {Sd(X)} }$  

$\therefore$  

$\hat{\beta_1} = 0.5 * {2Sd(X) \over Sd(X)} = 0.5 * 2 = 1$

__Answer 4:__ 1




# Question 5

Students were given two hard tests and scores were normalized to have empirical mean 0 and variance 1. The correlation between the scores on the two tests was 0.4. What would be the __expected score on Quiz 2__ for a student who had a normalized score of 1.5 on Quiz 1?

a) 0.16
b) 0.4
c) 1.0
d) __0.6__

## Solution

If sets are normalized, then $Y_i = X_i * Cor(X,Y)$

* $X_i$ = 1.5
* $Cor(X,Y)$ = 0.4
* $Y_i$ = ?

```{r}
1.5 * 0.4;
```

__Answer 5:__ 0.6


# Question 6

Consider the data given by the following

```{r}
x <- c(8.58, 10.46, 9.01, 9.64, 8.86)
```


What is the value of the __first measurement__ if x were normalized (to have mean 0 and variance 1)?

a) __-0.9719__
b) 8.58
c) 9.31
d) 8.86

## Solution

Normalize data by centering, then scaling: $Z_i= {{X_i−\bar{X}} \over S}$

```{r}
z <- ( (x - mean(x)) / sd(x) );

z;
```

__Answer 6:__ -0.9719



# Question 7

Consider the following data set (used for question 2 as well). What is the __intercept__ for fitting the model with x as the predictor and y as the outcome?
```{r}

x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42)
y <- c(1.39, 0.72, 1.55, 0.48, 1.19, -1.59, 1.23, -0.65, 1.49, 0.05)
```


a) __1.567__
b) 2.105
c) 1.252
d) -1.713

## Solution

Use  

$\hat{\beta_1} = Cor(Y,X) \cdot { {Sd(Y)} \over {Sd(X)} }$ and $\hat{\beta_0} = \bar{Y} - \hat{\beta_1}\bar{X}$

```{r}
ym <- mean(y); xm <- mean(x);
ys <- sd(y); xs <- sd(x);

b0 <- ym - (cor(x,y)*ys/xs) * xm; 

print(b0); print(lm(y~x));
```

Answers check out.

__Answer 7:__ 1.567




# Question 8

You know that both the predictor and response have mean 0. What can be said about the intercept when you fit a linear regression?

a) __It must be identically 0.__
b) Nothing about the intercept can be said from the information given.
c) It is undefined as you have to divide by zero.
d) It must be exactly one.

## Solution
By definition, the regression line goes through the mean values of both vectors

X and Y both having mean 0 means they have been centered by subtracting the means:
$$\tilde{X} = X_i - \bar{X} $$

The centering moves the origin $\{0,0\}$ to the middle of the plot, and the best linear fit line will go right through it.

```{r}
library("ggplot2");

x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42);
y <- c(1.39, 0.72, 1.55, 0.48, 1.19, -1.59, 1.23, -0.65, 1.49, 0.05);

xc <- x-mean(x); 
yc <- y-mean(y);

ggplot() + aes(xc,yc) + geom_point() + 
	geom_point(aes(x=0,y=0), color="red", size=4) +
	geom_abline(aes(intercept = coef(lm(yc~xc))[1], slope = coef(lm(yc~xc))[2]), color="red");

round(coef(lm(yc~xc)), 3);
```

__Answer 8:__ The intercept must be identically 0



# Question 9 

Consider the data given by

```{r}
x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42)
```

What value minimizes the sum of the squared distances between these points and itself?

a) __0.573__
b) 0.8
c) 0.36
d) 0.44

## Solution

For $\sum_{i=1}^{n}(X_i - \mu)^2$, the minimizing value is the sample mean: $\mu_{min} = \bar{X}$

```{r}
mean(x);
```

__Answer 9:__ 0.573



# Question 10

Let the _slope_ having fit Y as the outcome and X as the predictor be denoted as $\beta_1$. Let the slope from fitting X as the outcome and Y as the predictor be denoted as $\gamma_1$. Suppose that you divide $\beta_1$ by $\gamma_1$; in other words consider $\beta_1 / \gamma_1$. What is this ratio always equal to?

a) Var(Y)/Var(X)
b) 2SD(Y)/SD(X)
c) 1
d) Cor(Y,X)

## Solution

Use $\hat{\beta_1} = Cor(Y,X) \cdot { {Sd(Y)} \over {Sd(X)} }$

$$
{\beta_1 \over \gamma_1} =   Cor(Y,X) \cdot { {Sd(Y)}  \over {Sd(X)} } \div Cor(X,Y) \cdot { {Sd(X)} \over {Sd(Y)}}
$$

by $Cor(Y,X) = Cor(X,Y). . .$

$$
{\beta_1 \over \gamma_1} =  { {Sd(Y)} \over {Sd(X)} }  \div { {Sd(X)} \over {Sd(Y)} } 
= { {Sd(Y)} \over {Sd(X)}} \cdot { {Sd(Y)} \over {Sd(X)}}
$$

$$
\therefore {\beta_1 \over \gamma_1} = {Var(Y) \over Var{(X)}}
$$

Verify this
```{r}
x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42);
y <- c(1.39, 0.72, 1.55, 0.48, 1.19, -1.59, 1.23, -0.65, 1.49, 0.05);

# beta1
beta1 <- coef(lm(y~x))[2];

# gamma1
gamma1 <- coef(lm(x~y))[2];

beta1 / gamma1;

# answer a
var(y) / var(x); # <-- this is it

# answer b
2*sd(y)/sd(x)

# answer c
1;

# answer d
cor(x,y);
```

__Answer 10:__ var(y) / var(x)

